---
title: "Calcul of IBD in legacy datasets"
author: "MGH"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 4
    number_sections: true
    df_print: paged
---

# Set-up


```{r}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
setwd("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Geographic_sampling")
```

```{r, message=FALSE, warning=FALSE}
library(sf)
library(ggplot2)
library(ggpmisc)
library(tidyr)
library(tibble)
library(dplyr)
library(purrr)
library(units)
library("measurements")
library(geodist)
```


# Pairwise distance between populations

So here, I calculate the distance in km between pairs of populations. Everything is saved into a list with an element corresponding to a dataset (the pairwise Fst are also stored into a list, that's why!)

## Calculate the distances for datasets with GPS coordinates

### microsat datasets

```{r}
microsat_dist <- list.files("data_checked_ok/microsat/", full.names = TRUE)
```

```{r}
microsat_Dist_values_by_dataset <- list()

for (i in microsat_dist){
  #load coordinates of populations
  dataset <- read.csv(i,sep=";")
  dataset <- dataset %>% column_to_rownames("Pop") %>% as.matrix()
  
  #calculate the distance
  mat_dist <- geodist(dataset, measure = "geodesic") #measure in meter
  mat_dist <- mat_dist / 1000 #convert in km
  rownames(mat_dist) <- rownames(dataset)
  colnames(mat_dist) <- rownames(dataset)
  mat_dist
  
  #save that in a list
  name_dataset <- tools::file_path_sans_ext(basename(i))
  microsat_Dist_values_by_dataset[[name_dataset]] <- mat_dist
}


```

### SNPs datasets

```{r}
snps_dist <- list.files("data_checked_ok/snps/", full.names = TRUE)
```


```{r}
snps_Dist_values_by_dataset <- list()

for (i in snps_dist){
  #load coordinates of populations
  dataset <- read.csv(i,sep=";")
  dataset <- dataset %>% column_to_rownames("Pop_final") %>% as.matrix()
  
  #calculate the distance
  mat_dist <- geodist(dataset, measure = "geodesic") #measure in meter
  mat_dist <- mat_dist / 1000 #convert in km
  rownames(mat_dist) <- rownames(dataset)
  colnames(mat_dist) <- rownames(dataset)
  mat_dist
  
  #save that in a list
  name_dataset <- tools::file_path_sans_ext(basename(i))
  snps_Dist_values_by_dataset[[name_dataset]] <- mat_dist
}


```

## Add the distance matrix calculated by hand with ImageJ

```{r}
hand_calc_dist_mat <- list.files("data_checked_ok/dist_mat_hand_calc_with_ImageJ/", full.names = TRUE)
hand_calc_dist_mat

microsat_hand_calc_dist_mat <- c("data_checked_ok/dist_mat_hand_calc_with_ImageJ/Canis_lupus_Silva_2018_K11.txt","data_checked_ok/dist_mat_hand_calc_with_ImageJ/Canis_lupus_Szewczyk_2019.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Cypripedium_calceolus_Gargiulo_2023.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Mya_arenaria_DeNoia_2020.txt","data_checked_ok/dist_mat_hand_calc_with_ImageJ/Passer_domesticus_Baalsrud_2013_pop_1994.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Passer_domesticus_Baalsrud_2013_pop_2000.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Passer_domesticus_Baalsrud_2013_pop_2006.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Rana_arvalis_Cox_2024.txt")

snps_hand_calc_dist_mat <- c("data_checked_ok/dist_mat_hand_calc_with_ImageJ/Cypripedium_calceolus_Gargiulo_2023.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Passer_domesticus_Hagen_2020.txt","data_checked_ok/dist_mat_hand_calc_with_ImageJ/Sardina_pilchardus_Antoniou_2023.txt", "data_checked_ok/dist_mat_hand_calc_with_ImageJ/Symphonia_globulifera_Schmitt2021_Gargiulo2024.txt")
```


```{r}
for (i in microsat_hand_calc_dist_mat){
  mat <- as.matrix(read.table(i, header = TRUE, row.names = 1, sep = "\t", na.strings = ""))
  mat[upper.tri(mat)] <- t(mat)[upper.tri(mat)]
  
  #save that in a list
  name_dataset <- tools::file_path_sans_ext(basename(i))
  microsat_Dist_values_by_dataset[[name_dataset]] <- mat
  
  #check
  #print(name_dataset)
  #print(mat)
}

```

Let's add this in the SNPs list

```{r}
for (i in snps_hand_calc_dist_mat){
  mat <- as.matrix(read.table(i, header = TRUE, row.names = 1, sep = "\t", na.strings = ""))
  mat[upper.tri(mat)] <- t(mat)[upper.tri(mat)]
  
  #save that in a list
  name_dataset <- tools::file_path_sans_ext(basename(i))
  snps_Dist_values_by_dataset[[name_dataset]] <- mat
  
  #check
  #print(name_dataset)
  #print(mat)
}

```

## Check our final distance matrixes

```{r}
names(microsat_Dist_values_by_dataset)


names(snps_Dist_values_by_dataset)
```



and save these 2 lists

```{r}
save(microsat_Dist_values_by_dataset, file="microsat_Dist_values_by_dataset.Rdata")
save(snps_Dist_values_by_dataset, file="snps_Dist_values_by_dataset.Rdata")
```


# IBD

## Load the Fst pairwise
Load the pairwise matrices (already obtained from the Summary_statistics files)

```{r}
#SNPs
load("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v4/Summary_statistics/snps_Pairwise_values_by_dataset.Rdata")
names(snps_Pairwise_values_by_dataset)

#Microsat
load("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v2/FINAL_datasets/microsat_Pairwise_values_by_dataset.Rdata")
names(Pairwise_values_by_dataset_microsat)
```

Some adjustments --> improve the name of datasets

```{r}
#SNPs
names(snps_Pairwise_values_by_dataset) <- sub("^gen_", "", names(snps_Pairwise_values_by_dataset))
names(snps_Pairwise_values_by_dataset)

#microsat
names(Pairwise_values_by_dataset_microsat) <- sub("^genind_", "", names(Pairwise_values_by_dataset_microsat))
names(Pairwise_values_by_dataset_microsat)
```


## Table pairwise Fst/dist 

### Microsat

```{r}
microsat_datasets <- names(Pairwise_values_by_dataset_microsat)
```


```{r}
microsat_df_pop_2by2_Fst_dist <- list()

for (i in microsat_datasets){
  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(Pairwise_values_by_dataset_microsat[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(microsat_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #concat
  pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2), #this is to make sure that we have the 2 columns organized the same way as pop1-pop2 and not pop2-pop1
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))
  
  #store this dataframe in a list
  microsat_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
}


```

Checking if there are some problem
```{r}
has_NA <- sapply(microsat_df_pop_2by2_Fst_dist, anyNA)
names(microsat_df_pop_2by2_Fst_dist)[has_NA]
```

Cases with problems:

```{r}
i="Canis_lupus_Silva_2018_K11"

#turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(Pairwise_values_by_dataset_microsat[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
pairwise_fst_df

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(microsat_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
pairwise_dist_df
pairwise_dist_df$pop2 <- gsub("\\.", " ", pairwise_dist_df$pop2)
pairwise_dist_df

  #concat
pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2),
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))
  
  #store this dataframe in a list
  microsat_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
  microsat_df_pop_2by2_Fst_dist[[i]]
```

```{r}
i="Canis_lupus_Szewczyk_2019"

#turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(Pairwise_values_by_dataset_microsat[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
pairwise_fst_df

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(microsat_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
pairwise_dist_df
pairwise_dist_df$pop2 <- gsub("\\.", " ", pairwise_dist_df$pop2)
pairwise_dist_df

  #concat
pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2),
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))
  
  #store this dataframe in a list
  microsat_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
  microsat_df_pop_2by2_Fst_dist[[i]]
```

```{r}
i="Mya_arenaria_DeNoia_2020"

#turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(Pairwise_values_by_dataset_microsat[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  #pb comes because pb with pop name that was updated since
  new_pop_name <- c("K" = "ITA",
                    "Com_It" = "PLY",
                    "E"="TEX",
                    "Kiel_Germ"="KIE",
                    "SanAndr_UK"="SAN",
                    "H"="LIS",
                    "Bre_Fr"="BRE",
                    "LCt_Fr"="LCT")
  pairwise_fst_df <- pairwise_fst_df %>% 
  mutate(pop1 = recode(pop1, !!!new_pop_name, .default = pop1),
          pop2 = recode(pop2, !!!new_pop_name, .default = pop2))

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(microsat_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  pairwise_dist_df

  #concat
pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2),
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))
  
  #store this dataframe in a list
  microsat_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
  microsat_df_pop_2by2_Fst_dist[[i]]
```

```{r}
passer_datasets <- c("Passer_domesticus_Baalsrud_2013_pop_1994","Passer_domesticus_Baalsrud_2013_pop_2000","Passer_domesticus_Baalsrud_2013_pop_2006")

for (i in passer_datasets){
  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(Pairwise_values_by_dataset_microsat[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(microsat_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  pairwise_dist_df$pop2 <- gsub("\\.", "-", pairwise_dist_df$pop2)
  pairwise_dist_df
  
  #concat
  pairwise_fst_df  <- pairwise_fst_df %>% mutate(popA  = pmin(pop1, pop2),
                                                  popB = pmax(pop1, pop2)) %>% select(-pop1, -pop2)  
  pairwise_dist_df <- pairwise_dist_df  %>% mutate(popA  = pmin(pop1, pop2),  
                                                    popB = pmax(pop1, pop2))%>% select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_fst_df,pairwise_dist_df,by=c("popA", "popB")) #careful, order exchanged!!! because distance matrixes contains all the island while according the year, not exactly the same
  
  #store this dataframe in a list
  microsat_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
  microsat_df_pop_2by2_Fst_dist[[i]]
}

```

Now, look again at the problems:
```{r}
has_NA <- sapply(microsat_df_pop_2by2_Fst_dist, anyNA)
names(microsat_df_pop_2by2_Fst_dist)[has_NA]
```

If null --> then it's all good :)

### SNPs

```{r}
load("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v3/Geographic_measures/snps_Dist_values_by_dataset.Rdata")
names(snps_Dist_values_by_dataset) <- c("Barbastella_barbastellus_Razgour_2024",
                                        "Coenagrion_mercuriale_Leveque_2024",
                                        "Fagus_sylvatica_Milesi_2024",
                                        "Pinus_pinaster_Milesi_2024",
                                        "Pinus_sylvestris_Milesi_2024",
                                        "Raja_clavata_Marandel_2020",
                                        "Sardina_pilchardus_Antoniou_2023",
                                        "Passer_domesticus_Hagen_2020", 
                                        "Cypripedium_calceolus_Gargiulo_2023")
names(snps_Dist_values_by_dataset)
```

```{r}
snps_datasets <- names(snps_Pairwise_values_by_dataset)
snps_datasets

snps_datasets <- snps_datasets[snps_datasets != ("Symphonia_globulifera_Schmitt2021_Gargiulo2024")]
```


```{r}
snps_df_pop_2by2_Fst_dist <- list()

for (i in snps_datasets){
  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(snps_Pairwise_values_by_dataset[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(snps_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #concat
    pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2), #this is to make sure that we have the 2 columns organized the same way as pop1-pop2 and not pop2-pop1
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))

  #store this dataframe in a list
  snps_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
}


```

```{r}
has_NA <- sapply(snps_df_pop_2by2_Fst_dist, anyNA)
names(snps_df_pop_2by2_Fst_dist)[has_NA]
```


```{r}
i="Cypripedium_calceolus_Gargiulo_2023"

  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(snps_Pairwise_values_by_dataset[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  new_pop_name <- c("EK"="Korgessaare",
                    "EU"="Ussisoo")
  pairwise_fst_df <- pairwise_fst_df %>% 
  mutate(pop1 = recode(pop1, !!!new_pop_name, .default = pop1),
          pop2 = recode(pop2, !!!new_pop_name, .default = pop2))
  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(snps_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #concat
    pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2), #this is to make sure that we have the 2 columns organized the same way as pop1-pop2 and not pop2-pop1
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))

  #store this dataframe in a list
  snps_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
```


```{r}
i="Passer_domesticus_Hagen_2020"

  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(snps_Pairwise_values_by_dataset[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(snps_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  pairwise_dist_df
  
  #concat
  pairwise_fst_df  <- pairwise_fst_df %>% mutate(popA  = pmin(pop1, pop2),
                                                  popB = pmax(pop1, pop2)) %>% select(-pop1, -pop2)  
  pairwise_dist_df <- pairwise_dist_df  %>% mutate(popA  = pmin(pop1, pop2),  
                                                    popB = pmax(pop1, pop2))%>% select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_fst_df,pairwise_dist_df,by=c("popA", "popB")) #careful, order exchanged!!! because distance matrixes contains all the island while according the year, not exactly the same
  
  #store this dataframe in a list
  snps_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
  snps_df_pop_2by2_Fst_dist[[i]]
  
# 2 pop are missing from the mapping Leka and Vega so for now, let's look at the results without them
```


```{r}
i="Sardina_pilchardus_Antoniou_2023"

  #turn matrix fst in a dataframe
  pairwise_fst <- as.matrix(snps_Pairwise_values_by_dataset[[i]]$fst)
  mat_low <- lower.tri(pairwise_fst, diag = FALSE)
  pairwise_fst_df <- data.frame(pop1  = rownames(pairwise_fst)[row(pairwise_fst)[mat_low]],
                          pop2  = colnames(pairwise_fst)[col(pairwise_fst)[mat_low]],
                          Fst = as.numeric(pairwise_fst[mat_low]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)
  #pb comes because pb with pop name that was updated since
  new_pop_name <- c("1"="GSA1",
                    "10"="GSA10",
                    "11"="GSA11",
                    "19"="GSA19",
                    "5"="GSA5",
                    "6a"="GSA6a",
                    "6b"="GSA6b",
                    "6c"="GSA6c",
                    "9"="GSA9",
                    "GoL_1A"="GSA7a",
                    "GoL_2A"="GSA7b",
                    "Ob"="GoC")
  pairwise_fst_df <- pairwise_fst_df %>% 
  mutate(pop1 = recode(pop1, !!!new_pop_name, .default = pop1),
          pop2 = recode(pop2, !!!new_pop_name, .default = pop2))
  
  #turn matrix dist in a dataframe
  pairwise_dist <- as.matrix(snps_Dist_values_by_dataset[[i]])
  mat_low2 <- lower.tri(pairwise_dist, diag = FALSE)
  pairwise_dist_df <- data.frame(pop1  = rownames(pairwise_dist)[row(pairwise_dist)[mat_low2]],
                          pop2  = colnames(pairwise_dist)[col(pairwise_dist)[mat_low2]],
                          Dist_km = as.numeric(pairwise_dist[mat_low2]),
                          row.names = NULL,
                          stringsAsFactors = FALSE)

  #concat
    pairwise_fst_df  <- pairwise_fst_df %>%
  mutate(popA  = pmin(pop1, pop2), #this is to make sure that we have the 2 columns organized the same way as pop1-pop2 and not pop2-pop1
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
pairwise_dist_df <- pairwise_dist_df  %>%
  mutate(popA  = pmin(pop1, pop2),  
         popB = pmax(pop1, pop2))%>%             
  select(-pop1, -pop2)  
  Fst_dist_df <- left_join(pairwise_dist_df,pairwise_fst_df,by=c("popA", "popB"))

  #store this dataframe in a list
  snps_df_pop_2by2_Fst_dist[[i]] <- Fst_dist_df
```

Let's check again if it's okay (only Passer domesticus Hagen should be here)
```{r}
has_NA <- sapply(snps_df_pop_2by2_Fst_dist, anyNA)
names(snps_df_pop_2by2_Fst_dist)[has_NA]
```

## Plot the IBD: Rousset stepping-stone model

As in Milesi et al. 2024 --> following the Rousset approach for a two-dimensional stepping-stone model: 
Fst/(1-Fst)=b*log(dist)+a+er

```{r}
for (i in names(microsat_df_pop_2by2_Fst_dist)){
  df <- microsat_df_pop_2by2_Fst_dist[[i]]
  g <- ggplot(df,aes(x=log(Dist_km),y=(Fst/(1-Fst))))+geom_point()+ 
    geom_smooth(method = "lm", color = "blue") +
    stat_poly_eq(
    formula = y ~ x,
    aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~~~")),
    colour = "blue") +
  ggtitle(i)+
    ylim(0,0.8)+
  theme_minimal()
  print(g)
}


```


```{r}
for (i in names(snps_df_pop_2by2_Fst_dist)){
  df <- snps_df_pop_2by2_Fst_dist[[i]]
  g <- ggplot(df,aes(x=log(Dist_km),y=(Fst/(1-Fst))))+geom_point()+ 
    geom_smooth(method = "lm", color = "blue") +
    stat_poly_eq(
    formula = y ~ x,
    aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~~~")),
    colour = "blue") +
  ggtitle(i)+
    ylim(0,0.8)+
  theme_minimal()
  print(g)
}


```


## Gather the regression coeff of IBD in a tab

As in Milesi et al. 2024 --> following the Rousset approach for a two-dimensional stepping-stone model: 
Fst/(1-Fst)=b*log(dist)+a+er

```{r}
#replace negative Fst by 0
microsat_df_pop_2by2_Fst_dist_2 <- map(microsat_df_pop_2by2_Fst_dist, ~ .x %>% mutate(Fst = pmax(Fst, 0)))
snps_df_pop_2by2_Fst_dist_2 <- map(snps_df_pop_2by2_Fst_dist, ~ .x %>% mutate(Fst = pmax(Fst, 0)))

  
IBD_coeff_Rousset <- matrix(ncol=8,nrow=0)
colnames(IBD_coeff_Rousset) <- c("Dataset","Marker_type","a_intercept","pval_a_intercept","b_slope","pval_b_slope","R2","R2_adjusted")

## Microsat
Marker_type="microsat"
for (i in 1:length(microsat_df_pop_2by2_Fst_dist_2)){
  
  dat=microsat_df_pop_2by2_Fst_dist_2[[i]]
  #sometimes, distance are very small and interpreted as zero
  dat <- dat %>% mutate(Dist_km=case_when(Dist_km == 0 ~ 0.00000001,
                                          TRUE ~ Dist_km))
  
  
  #Linear model: Fst/(1-Fst)=b*log(dist)+a+er
  mod <- lm((Fst/(1 - Fst)) ~ log(Dist_km), data = dat)
  
  IBD_coeff_Rousset <- rbind(IBD_coeff_Rousset, 
                             c(names(microsat_df_pop_2by2_Fst_dist_2)[[i]],#name of the dataset
                               Marker_type,
                               summary(mod)$coefficients[1,1],#intercept coeff
                               summary(mod)$coefficients[1,4],#intercept p-value
                               ifelse(nrow(summary(mod)$coefficients) >= 2,summary(mod)$coefficients[2, 1],"NA"),#slope coeff
                               ifelse(nrow(summary(mod)$coefficients) >= 2,summary(mod)$coefficients[2, 4],"NA"),#slope p-value
                               summary(mod)$r.squared,#R2
                               summary(mod)$adj.r.squared #R2-adjusted
                               ))

}


## SNPs
Marker_type="snps"
for (i in 1:length(snps_df_pop_2by2_Fst_dist)){
  
  dat=snps_df_pop_2by2_Fst_dist[[i]]
  #sometimes, distance are very small and interpreted as zero
  dat <- dat %>% mutate(Dist_km=case_when(Dist_km == 0 ~ 0.00000001,
                                          TRUE ~ Dist_km))
  
  
  #Linear model: Fst/(1-Fst)=b*log(dist)+a+er
  mod <- lm((Fst/(1 - Fst)) ~ log(Dist_km), data = dat)
  
  IBD_coeff_Rousset <- rbind(IBD_coeff_Rousset, 
                             c(names(snps_df_pop_2by2_Fst_dist)[[i]],#name of the dataset
                               Marker_type,
                               summary(mod)$coefficients[1,1],#intercept coeff
                               summary(mod)$coefficients[1,4],#intercept p-value
                               ifelse(nrow(summary(mod)$coefficients) >= 2,summary(mod)$coefficients[2, 1],"NA"),#slope coeff
                               ifelse(nrow(summary(mod)$coefficients) >= 2,summary(mod)$coefficients[2, 4],"NA"),#slope p-value
                               summary(mod)$r.squared,#R2
                               summary(mod)$adj.r.squared #R2-adjusted
                               ))

  }
```

Save this table
```{r}
# add a column with IBD_slope x 10^(-3)
IBD_coeff_Rousset <- as.data.frame(IBD_coeff_Rousset)
IBD_coeff_Rousset <- IBD_coeff_Rousset %>% mutate(IBD_slope_10_3=as.numeric(b_slope)*(10^(3)))

#save
write.csv(IBD_coeff_Rousset, file="IBD_snps_microsat_Rousset_reg.csv")
```


## Graphical summary

Look at the values
```{r}
IBD_coeff_Rousset$Dataset_Marker_type <- paste0(IBD_coeff_Rousset$Dataset,"-",IBD_coeff_Rousset$Marker_type)
IBD_coeff_Rousset$R2_adjusted <- as.numeric(IBD_coeff_Rousset$R2_adjusted)
IBD_coeff_Rousset$R2 <- as.numeric(IBD_coeff_Rousset$R2)

IBD_coeff_Rousset %>% 
  arrange(Dataset_Marker_type) %>%
  select(Dataset_Marker_type, IBD_slope_10_3, R2) %>%
  mutate(IBD_slope_10_3 = round(IBD_slope_10_3, 2),
         R2 = round(R2, 3)) %>%
  print()
```

```{r}

IBD_coeff_Rousset %>% ggplot(aes(x=reorder(Dataset_Marker_type, IBD_slope_10_3),y=IBD_slope_10_3))+
                            geom_point()+
                            theme(axis.text.x = element_text(angle=90, hjust=1))+
                            xlab("Dataset")+ylab("Slope x 10^(-3)")+
                            ggtitle("Slope of regression for Fst/(1-Fst)=log(Dist)")

IBD_coeff_Rousset %>% ggplot(aes(x=R2,y=IBD_slope_10_3,fill=Dataset_Marker_type, color=Dataset_Marker_type))+
                            geom_point()+
                            theme(axis.text.x = element_text(angle=90, hjust=1))+
                            xlab("Quality of regression (R²)")+ylab("Slope x 10^(-3)")+
                            ggtitle("Regression of Fst/(1-Fst)=log(Dist)")

IBD_coeff_Rousset %>% ggplot(aes(x=R2,y=IBD_slope_10_3,fill=Dataset_Marker_type, color=Dataset_Marker_type))+
                            geom_point()+
                            theme(axis.text.x = element_text(angle=90, hjust=1))+
                            theme(legend.position="none")+
                            xlab("Quality of regression (R²)")+ylab("Slope x 10^(-3)")+
                            ggtitle("Regression for Fst/(1-Fst)=log(Dist)")


```

Most of the regression do not really have a good quality as shown with low values of R². 
Only 5 datasets with R² above 0.5.

Varying values of the slopes according the species. For the comparison of SNPs vs microsat, we only have Passer domesticus, seems congruent. 


