---
title: "SNPs_filtering_on_splitted_pop_full"
author: "MGH"
date: "2024-11-18"
output: html_document
---


Objective: code for ensuring good quality of SNP datasets, here automatized for different datasets, with a scope of estimating NeLD. A complementary doc with the part of code in bash is also be available. 

See these reference that we used to set up the different thresholds of filtering: 
- https://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/#ex1
- Hemstrom et al. 2024, Next-Generation Data Filtering in the Genomics Era 
- Waples 2024, Practical Application of the Linkage Disequilibrium Method for Estimating Contemporary Effective Population Size: A Review 

# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```

```{r}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
setwd("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v2")
```

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

library(purrr)

library(vcfR)
library(adegenet)
```


# 1. Get the mean read depth for the raw vcf files


```{r}
files_list <- list.files("raw_data/snps/", pattern = "\\.vcf$", full.names = TRUE)

basename_list <- sub("^raw_data/snps/", "", files_list)
basename_list <- sub("\\.vcf", "", basename_list)

stats_DP <- matrix(ncol=3,nrow=0)

pdf("raw_vcf_DP_hist.pdf")

for (i in basename_list){

  load(paste0("raw_data/snps/R_objects/vcfR_", i,".Rdata"))
  assign("vcf_file",get(ls()[startsWith(ls(), "vcf_")]))
  
  recode <- vcf_file %>% 
      extract_gt_tidy(format_fields = c("GT","DP", "GQ")) %>% 
      mutate(gt_GT = dplyr::recode(gt_GT, 
                               '0|1' = '0/1', '1|0' = '1/0', 
                               '1|1' = '1/1', '0|0' = '0/0'),
          Indiv = Indiv)

  # Load info in a table
  mean_DP <- mean(recode$gt_DP,na.rm=T)
  max_DP <- max(recode$gt_DP,na.rm=T)

  stats_DP <- rbind(stats_DP,c(vcf_file= i,
                               mean_DP= mean_DP,
                               max_DP= max_DP))
  
  #if info on DP in the vcf, plot the histogram
  if (!is.na(mean_DP)){
    print(hist(recode$gt_DP, main=i))
  } 
  
  
  #cleaning
  rm(list=ls()[startsWith(ls(), "vcf_")])
  rm(recode,mean_DP,max_DP)
} 

dev.off()

stats_DP <- as.data.frame(stats_DP)
stats_DP$mean_DP <- as.numeric(stats_DP$mean_DP)
stats_DP$max_DP <- as.numeric(stats_DP$max_DP)
stats_DP$twice_mean_DP <- round(stats_DP$mean_DP*2,2)

write.table(stats_DP, row.names = FALSE, quote=FALSE, col.names = FALSE,
          file="mean_RD_raw_vcf_full.txt")
```



# 2. Filtering

Launch the code "Filtering_snps_full.sh"

It's possible to change the parameters of filtering, see in the code file, at the top of the document.


# 3. Graphical representations

```{r}
files_list <- list.files("raw_data/snps/", pattern = "\\.vcf$", full.names = TRUE)
basename_list <- sub("^raw_data/snps/", "", files_list)
basename_list <- sub("\\.vcf", "", basename_list)
```


## Before filtering on missing data

Graph after quality filtering to check for missing data, per SNPs and per individuals

```{r}
#set-up the function
graph_representation_missingness <- function(i){
path="filtered_data/MIN_DP_6_MD_SNPs_0.2_Ind_0.2_full_vcf/" #change that if you have set different parameters
  
 #variant missingness
var_miss <- read_delim(paste0(path,"intermed_files/",i,"_geno_qual.lmiss"), delim = "\t",
                       col_names = c("chr", "pos", "nchr", "nfiltered", "nmiss", "fmiss"), skip = 1)
g <- ggplot(var_miss, aes(fmiss)) + geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) + theme_light() +
          xlim(0,1)+
          ggtitle(paste(i,"\n %MD per SNPs across ind"))
print(g)
ggsave(g,file=paste0(path,"graphics/before_MD_filtering/per_snps_",i,".pdf"))
summary(var_miss$fmiss)

#individual missingness
ind_miss  <- read_delim(paste0(path,"intermed_files/",i,"_geno_qual.imiss"), delim = "\t",
                        col_names = c("ind", "ndata", "nfiltered", "nmiss", "fmiss"), skip = 1)
q <- ggplot(ind_miss, aes(fmiss)) + geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3) + theme_light() +
          xlim(0,1)+
          ggtitle(paste(i,"\n %MD per ind across SNPs"))
print(q)
ggsave(q,file=paste0(path,"graphics/before_MD_filtering/per_ind_",i,".pdf"))
summary(ind_miss$fmiss) 
}
```

Let's run the code to get our graph, go check in folder "graphics"
```{r}
#apply it to all our files (with purr package to avoid loop for)
walk(basename_list,graph_representation_missingness)
```



## After filtering: MAF profile

After all the steps of filtering, plotting of the MAF in our final datasets

```{r}
graph_MAF <- function(i){
path="filtered_data/MIN_DP_6_MD_SNPs_0.2_Ind_0.2_full_vcf/"
  
#MAF
var_freq <- read_delim(paste0(path,"intermed_files/",i,"_MD_snps_ind.frq"), delim = "\t",
                       col_names = c("chr", "pos", "nalleles", "nchr", "a1", "a2"), skip = 1)
  var_freq$maf <- var_freq %>% select(a1, a2) %>% apply(1, function(z) min(z))
  p <- ggplot(var_freq, aes(maf)) + geom_histogram(bins=200, binwidth = 0.01, colour="black", fill="blue") + theme_light() +
    ggtitle(paste("MAF - ",i))
  print(p)
ggsave(p,file=paste0(path,"graphics/after_MD_filtering/MAF_",i,".pdf"))

#MAC
var_count <- read_delim(paste0(path,"intermed_files/",i,"_MD_snps_ind.frq.count"), delim = "\t",
                       col_names = c("chr", "pos", "nalleles", "nchr", "a1", "a2"), skip = 1)
  var_count <- separate(var_count, a1, c("all_a1","a1"), sep=":")
  var_count <- separate(var_count, a2, c("all_a2","a2"), sep=":")
  var_count$a1 <- as.numeric(var_count$a1)
  var_count$a2 <- as.numeric(var_count$a2)
  var_count$mac <- var_count %>% select(a1, a2) %>% apply(1, function(z) min(z))
  
  p <- ggplot(var_count, aes(mac)) + geom_histogram(bins=200, binwidth = 0.01, colour="black", fill="blue") + theme_light() +
    ggtitle(paste("MAC - ",i))
  print(p)
ggsave(p,file=paste0(path,"graphics/after_MD_filtering/MAC_",i,".pdf"))


}
```

```{r}
#apply it to all our files (with purr package to avoid loop for)
walk(basename_list,graph_MAF)
```


