---
title: "NeEstims_withEstim"
author: "MGH"
date: "2025-06-17"
output: html_document
---

Here, the purpose of this code is to read the results files from Estim (that I previously ran by hand, as the software has a windows graphical interface). And then gather the results in a single tab with all the summary statistics (F,n,Ne,m and their confidence intervals). 

# Set-up

```{r}
rm(list=ls())
setwd("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/NeEstimations")
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
```


# Gather the files in a tab
 Idea is to adapt the code I wrote for the extraction of results from NeEstimator. 

```{r}
#function to extract values in R
extract_values <- function(line) {
  values <- unlist(strsplit(line, "\\s+"))
  values <- values[values != ""]
}

extract_numeric_values <- function(x) {
  matches <- regmatches(x, gregexpr("\\d+,\\d+", x))
  as.numeric(gsub(",", ".", unlist(matches)))
}

handle_negatives <- function(tokens) {
  result <- c()
  i <- 1
  while (i <= length(tokens)) {
    
    if (tokens[i] == "-") {
      if (i < length(tokens) && grepl("^\\d+,\\d+$", tokens[i + 1])) {
        result <- c(result, paste0("-", tokens[i + 1]))
        i <- i + 2
      } else {
        result <- c(result, NA)
        i <- i + 1
      }
    } else if (grepl("-", tokens[i], fixed = TRUE) && !grepl("^\\-?\\d+,\\d+,?$", tokens[i])) {
      
      result <- c(result, NA)
      i <- i + 1
    } else {
      result <- c(result, tokens[i])
      i <- i + 1
    }
  }
  result
}

extract_numeric_values <- function(x) {
  # On split la ligne par espace
  #tokens <- unlist(strsplit(x, "\\s+"))
  tokens <- x
  tokens <- tokens[!tokens %in% c("","->")]
  
  # On nettoie les crochets
  tokens <- gsub("\\[|\\]", "", tokens)
  
  # About the "-" -->
  tokens <- handle_negatives(tokens)
  
  #Transfo of commas
  tokens <- gsub(",$", "", tokens)
  tokens <- gsub(",", ".", tokens)
}


```

Set up
```{r}
#Table where every info is going to be kept
results_Estim <- matrix(NA, ncol=14, nrow=0)

#Path to the files
result_files <- list.files("with_Estim/results", full.names = TRUE)
```


```{r}
for (i in result_files) {
      
      #######################################
      #Set up
      text_file=i
      lines <- readLines(text_file)
      #Lines with interesting data
      start_index <- grep("Population", lines)
      end_index <- grep("^$", lines)
      
      #######################################
      #Get interesting lines from tab 1
      table_lines_2stats <- lines[(start_index[1] + 2):(end_index[2] - 2)]
      
      #Split this full lines into columns
      table_lines_2stats <- lapply(table_lines_2stats, extract_values)
    
      #Get the numeric values
      table_lines_2stats <- lapply(table_lines_2stats, extract_numeric_values)
      
      #A data frame
      df_2stats <- as.data.frame(do.call(rbind, table_lines_2stats),header=FALSE)
      
      #######################################
      #Get interesting lines from tab 2
      table_lines_Nem <- lines[(start_index[2] + 3):(end_index[3] - 2)]
      
      #Split this full lines into columns
      table_lines_Nem <- lapply(table_lines_Nem, extract_values)
      #Get the numeric values
      table_lines_Nem <- lapply(table_lines_Nem, extract_numeric_values)
      
      #A data frame
      df_Nem <- as.data.frame(do.call(rbind, table_lines_Nem))
      
      ############################################
      # FINAL DATAFRAME
      results_Estim <- rbind(results_Estim,cbind(sub("\\.txt$", "",basename(i)), #Dataset
                                                 df_2stats, # first col, pop index then F and etaS 
                                                 df_Nem[,-1])) #Ne and m 
      
      #cleaning
      rm(end_index,start_index,lines,text_file,table_lines_2stats,table_lines_Nem,df_2stats,df_Nem)

}

colnames(results_Estim) <- c("Dataset","Index_pop","F","F_CI_down","F_CI_up","etaS","etaS_CI_down","etaS_CI_up",
                             "Ne","Ne_CI_down","Ne_CI_up","m","m_CI_down","m_CI_up")

```


Yeeeah! it worked!

What about adding the population name...? 



Let's save this files of results in a tab
```{r}
write.csv(results_Estim,file="with_Estim/summary_results_Ne_withEstim.csv",
          row.names=FALSE)
```



