---
title: "Estimation of Ne with ONeSAMP 3 - only for SNPs"
author: "MGH"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 4
    number_sections: true
    df_print: paged
---

Here, the purpose of this code is to estimate Ne with the software ONeSAMP 3.0 (Hong et al. 2023). 

See here for more details on the software: https://github.com/AaronHong1024/ONeSAMP_3

This method of estimation is based on summary statistics and Bayesian computation with linear regression. It works only with SNPs datasets (not microsatellites). 

First of all, I had to:
- reencoded my genepop files for them to be read by ONeSAMP3 (only 2 digits per allele, individual name as numbers)
- ran ONeSAMP 3.0 on the cluster, as the software is computationally intensive, see script_cluster_ONESAMP_subsamp_1000snps.sh

And then I want to gather the results in a single table with all the summary statistics 

Example of a result file : 

WARNING:main: No filename provided.  Using ridge.joblib
Below are the mean, median, and 95 credible limits for the posterior distribution of the effective population size from OneSamp
min        max        mean        median      lower95CL   upper95CL
63.98      243.87      96.61      97.41      74.73      152.36
--- 45649.141025304794 seconds ---

--> I think it's important to report the time of running, because it's very long

# Set-up

```{r}
rm(list=ls())
setwd("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/NeEstimations")
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(purrr)
```

# Renencoding genepop files for ONeSAMP 3

I need reencode the genepop files to be read by ONeSAMP 3 (otherwise it does not work). The things to rencode are the following:
- have alleles coded in 2 characters instead of 3,
- the individuals name need to be coded as numbers, from 1, 2, 3 to n... 

## For subsamp 1000 snps

```{r}
list_snps_file <- 
  list.files("genpop_files/subsamp_1000snps/", full.names = TRUE)
nb_snps=1000
#list_snps_file <- list_snps_file[1:311]
```


```{r, eval=FALSE}
for (i in list_snps_file){

  #let's read the file by lines
  lines <- readLines(i)

  #find pop indice
  pop_indices <- grep("^POP", lines, ignore.case = TRUE)
  if(length(pop_indices) != 1){
    stop("Not a single pop in this dataset, check please")
  }

  #keep the header
  header <- lines[1:(pop_indices)]
  
  ########################################################
  #create a dataframe with genotypes
  data_lines <- lines[(pop_indices+1):length(lines)]

  geno_tab <- matrix(ncol=(nb_snps+1), nrow = 0)

  for (j in seq_along(data_lines)){
    current_line <- trimws(data_lines[j])
    parts <- strsplit(current_line, ",")[[1]]
    #ind name
    ind <- trimws(parts[1])
    #genotypes line = 1 by locus
    full_geno <- trimws(parts[2])
    full_geno <- unlist(strsplit(full_geno, "\\s+"))
  
    n_loci <- length(full_geno)
    if(all(n_loci == nb_snps)){
    # add this line of genotype in the tab of genetic data we're building
      geno_tab <- rbind(geno_tab,c(ind,full_geno))
    }else{
      stop("Not the same nb of loci between indviduals, weird so check please")
    }
  }

  #transform it into a dataframe
  geno_tab <- as.data.frame(geno_tab)
  snps_name <- paste0("loc_",seq_along(1:nb_snps))
  colnames(geno_tab) <- c("ind", snps_name)

  #Let's split into 2 columns 
  geno_df_recoded <- geno_tab %>%
    mutate(across(-ind,
          list(all1 = ~ substr(.x, 1, 3),
              all2 = ~ substr(.x, 4, 6)),
          .names = "{.col}_{.fn}")) %>%
      select(ind, ends_with("all1"), ends_with("all2"))
  ########################################################

  # create a list with the unique alleles of each locus
  alleles_list <- map(snps_name, function(snp) {
    allele_columns <- geno_df_recoded %>% select(starts_with(snp))
    unique(unlist(allele_columns))
  })
  names(alleles_list) <- snps_name

  #let's add the info for recoding
  alleles_list_df <- map(alleles_list, function(allele_vec) {
    # keep sequence of alleles
    alleles_unique <- unique(allele_vec)
    #remove missing alleles (specific case)
    non_missing <- alleles_unique[alleles_unique != "000"]
    #sequence to attribute encoded alleles with 2 caracters
    codes <- sprintf("%02d", seq_along(non_missing))
    # corresponding table
    mapping <- setNames(codes, non_missing)
    #specific case of missing data
    if ("000" %in% alleles_unique) {
      mapping["000"] <- "00"
    }
    #a dataframe with 2 columns, original=encoding with 3 caracters/recoded=encoding with 2 caracters
    data.frame(original = names(mapping), recoded = as.vector(mapping), 
             stringsAsFactors = FALSE)
  })

  ########################################################
  #reencode the alleles in a dataframe
  geno_df_final <- data.frame(ind = seq_along(1:nrow(geno_tab)), stringsAsFactors = FALSE)
  for (snp in snps_name){
    new_col <- geno_df_recoded %>% select(starts_with(paste0(snp,"_")))
    alleles <- alleles_list_df[[snp]]
  
    new_col <- new_col %>% mutate(across(everything(), 
                                       ~ recode(., !!! setNames(alleles$recoded, alleles$original))))
  
    geno_df_final <- bind_cols(geno_df_final,new_col)
  }

  ###########################################################################
  #now let's return to the matrix format and reput everything in a genepop format
  genepop_lines <- apply(geno_df_final, 1, function(x) {
    x <- as.character(x)
    id <- x[1]
    alleles <- x[-1]
    # paste alleles 2 by 2
    pairs <- paste0(
      alleles[seq(1, length(alleles), by = 2)],
      alleles[seq(2, length(alleles), by = 2)])
    # put the line in the good shape, with info split by spaces (2 spaces after the comma)
    paste(id, ", ", paste(pairs, collapse = " "))
  })

  #add the header
  genepop_lines <- c(header,genepop_lines)

  #save this newly renecoded genepop file
  name_file <-  sub("^.*/(.*)\\.[^.]+$", "\\1", i)
  writeLines(genepop_lines, paste0("with_ONeSAMP/reencoded_genepop_files/subsamp_1000snps/",name_file,".txt"))
  
  ########################################################
  
  #check that loop was okay
  cat("which pop are we?",name_file)
  cat("dim tab before reencoding",dim(geno_tab))
  cat("dim tab after reencoding",dim(geno_df_final))
  cat("Nb snps in the loop should be",nb_snps,", here = ",snp)
  
  #cleaning
  rm(current_line,data_lines,full_geno,genepop_lines,header,ind,lines,n_loci,parts,pop_indices,snp,snps_name,j)
  rm(alleles,alleles_list,alleles_list_df,geno_df_final,geno_df_recoded,geno_tab, new_col)
}
```


## SNPs less than 1000

```{r}
list_full_snps_file <- list.files("genpop_files/snps_full/", full.names = TRUE)

list_few_snps_file <- list_full_snps_file[grep("Ursus_arctos|Alytes_obstetricans",list_full_snps_file)]
print(list_few_snps_file)
```


```{r, eval=FALSE}
for (i in list_few_snps_file){

  #let's read the file by lines
  lines <- readLines(i)

  #find pop indice
  pop_indices <- grep("^POP", lines, ignore.case = TRUE)
  if(length(pop_indices) != 1){
    stop("Not a single pop in this dataset, check please")
  }

  #keep the header
  header <- lines[1:(pop_indices)]
  #two cases of header according the way of converting the file to genepop 
  if (length(header) > 3){
    #case of use PGDSpider
     nb_snps <- (length(header)-2) #header minus empty first line and last line with "POP"
  } else{
    # eg case of use of package graph4lg
    parts <- trimws(unlist(strsplit(header[2], ",")))
    nb_snps <- length(parts)
  }
  
  
  ########################################################
  #create a dataframe with genotypes
  data_lines <- lines[(pop_indices+1):length(lines)]

  geno_tab <- matrix(ncol=(nb_snps+1), nrow = 0)

  for (j in seq_along(data_lines)){
    current_line <- trimws(data_lines[j])
    parts <- strsplit(current_line, ",")[[1]]
    #ind name
    ind <- trimws(parts[1])
    #genotypes line = 1 by locus
    full_geno <- trimws(parts[2])
    full_geno <- unlist(strsplit(full_geno, "\\s+"))
  
    n_loci <- length(full_geno)
    if(all(n_loci == nb_snps)){
    # add this line of genotype in the tab of genetic data we're building
      geno_tab <- rbind(geno_tab,c(ind,full_geno))
    }else{
      stop("Not the same nb of loci between indviduals, weird so check please")
    }
  }

  #transform it into a dataframe
  geno_tab <- as.data.frame(geno_tab)
  snps_name <- paste0("loc_",seq_along(1:nb_snps))
  colnames(geno_tab) <- c("ind", snps_name)

  #Let's split into 2 columns 
  geno_df_recoded <- geno_tab %>%
    mutate(across(-ind,
          list(all1 = ~ substr(.x, 1, 3),
              all2 = ~ substr(.x, 4, 6)),
          .names = "{.col}_{.fn}")) %>%
      select(ind, ends_with("all1"), ends_with("all2"))
  ########################################################

  # create a list with the unique alleles of each locus
  alleles_list <- map(snps_name, function(snp) {
    allele_columns <- geno_df_recoded %>% select(starts_with(snp))
    unique(unlist(allele_columns))
  })
  names(alleles_list) <- snps_name

  #let's add the info for recoding
  alleles_list_df <- map(alleles_list, function(allele_vec) {
    # keep sequence of alleles
    alleles_unique <- unique(allele_vec)
    #remove missing alleles (specific case)
    non_missing <- alleles_unique[alleles_unique != "000"]
    #sequence to attribute encoded alleles with 2 caracters
    codes <- sprintf("%02d", seq_along(non_missing))
    # corresponding table
    mapping <- setNames(codes, non_missing)
    #specific case of missing data
    if ("000" %in% alleles_unique) {
      mapping["000"] <- "00"
    }
    #a dataframe with 2 columns, original=encoding with 3 caracters/recoded=encoding with 2 caracters
    data.frame(original = names(mapping), recoded = as.vector(mapping), 
             stringsAsFactors = FALSE)
  })

  ########################################################
  #reencode the alleles in a dataframe
  geno_df_final <- data.frame(ind = seq_along(1:nrow(geno_tab)), stringsAsFactors = FALSE)
  for (snp in snps_name){
    new_col <- geno_df_recoded %>% select(starts_with(paste0(snp,"_")))
    alleles <- alleles_list_df[[snp]]
  
    new_col <- new_col %>% mutate(across(everything(), 
                                       ~ recode(., !!! setNames(alleles$recoded, alleles$original))))
  
    geno_df_final <- bind_cols(geno_df_final,new_col)
  }

  ###########################################################################
  #now let's return to the matrix format and reput everything in a genepop format
  genepop_lines <- apply(geno_df_final, 1, function(x) {
    x <- as.character(x)
    id <- x[1]
    alleles <- x[-1]
    # paste alleles 2 by 2
    pairs <- paste0(
      alleles[seq(1, length(alleles), by = 2)],
      alleles[seq(2, length(alleles), by = 2)])
    # put the line in the good shape, with info split by spaces (2 spaces after the comma)
    paste(id, ", ", paste(pairs, collapse = " "))
  })

  #add the header
  genepop_lines <- c(header,genepop_lines)

  #save this newly renecoded genepop file
  name_file <-  sub("^.*/(.*)\\.[^.]+$", "\\1", i)
  writeLines(genepop_lines, paste0("with_ONeSAMP/reencoded_genepop_files/fullsnps/",name_file,".txt"))
  
  ########################################################
  
  #check that loop was okay
  cat("which pop are we?",name_file)
  cat("dim tab before reencoding",dim(geno_tab))
  cat("dim tab after reencoding",dim(geno_df_final))
  cat("Nb snps in the loop should be",nb_snps,", here = ",snp)
  
  #cleaning
  rm(current_line,data_lines,full_geno,genepop_lines,header,ind,lines,n_loci,parts,pop_indices,snp,snps_name,j)
  rm(alleles,alleles_list,alleles_list_df,geno_df_final,geno_df_recoded,geno_tab, new_col)
}
```


# Run ONeSAMP 3 (on the cluster)

For more details, you can see on script_cluster_ONESAMP_subsamp_1000snps.sh

NB: when run on several dataset at the same time within a single folder containing the program, there is a problem with temporary files created. So make sure to clone several times the program in different folders if you want to run it in parallel on several files. 

Here are the parameters we used:
- m 0.000001 - Minimum allele frequency, to be considered as zero, as we already filtered our data for rare alleles
- s 50000 - Number of ONeSAMP trials, as big as possible to add the more iteration possible for more robust result with this Bayesian framework
- lNe 5 -uNe 400 - Range of Ne, wider allowed (look in the code for upper range)
- lT 0.00005 -uT 0.05 - Range of Theta (Theta=Ne*mu), wide also and realistic for SNPs markers

We also tried to run ONeSAMP 3, on datasets with expected Ne bigger than 400 (above 1000) and we got unrealistic outlier values (that could be related to -Infinite or Infinite). Thus, I only used the program on the following datasets:
- Alytes obstetricans 
- Barbastella barbastellus (although one of the 2 population as an NeLD ~ 1000 with ldne) - Colonies in Iberia and Sussex 
- Cypripedium calceolus - Estonia populations
- Passer domesticus - metapopulation inslands in Norway
- Ursus arctos - Fennoscandinavian population


# Identify the files that did not run correctly

I know that there were some problems with the running of some files 

see for example

```{r}
readLines("with_ONeSAMP/results_subsamp_1000snps_07_25/Barbastella_barbastellus_Razgour_2024_Iberian_col_1000snps_subset_5_output.txt")
```

So to list them, we just have to count the number of lines (if longer than 6 then there were a problem)

```{r}
results_ONeSAMP <- list.files("with_ONeSAMP/results_subsamp_1000snps_07_25/", full.names = TRUE)
results_ONeSAMP <- results_ONeSAMP[results_ONeSAMP != "with_ONeSAMP/results_subsamp_1000snps_07_25/results_28_07"]

problem_running_files <- c()

for (i in results_ONeSAMP){
  file_result <- readLines(i)
  
  if (length(file_result) > 6){
    problem_running_files <- c(problem_running_files, i)
  }
}

problem_running_files
length(problem_running_files)
```

ok so 41/62 files with a problem... almost all files of Passer domesticus, most of Barbastella also... This was due to the problem of dependencies/temporary files created in background when using a single folder containing the program and several files.


# Gather results in a tab

```{r}
#function to extract values in R
extract_values <- function(line) {
  values <- unlist(strsplit(line, "\\s+"))
  values <- values[values != ""]
}
```


```{r}

#Table where every info is going to be kept
Ne_ONeSAMP <- matrix(NA, ncol=11, nrow=0)
colnames(Ne_ONeSAMP) <- c("Dataset","Pop","Subset","Pb","min","max","mean","median","lower95CL","upper95CL","computation_time")

#Path to the files
for (i in results_ONeSAMP) {

      text_file=i
      #info about the dataset
      parts <- strsplit(basename(i), "_")[[1]]
      dataset <- paste(parts[1:4], collapse = "_")
      pop <- paste(parts[5:(length(parts)-4)], collapse = "_")
      subset <- parts[length(parts)-1]
      
      #result tab
      lines <- readLines(text_file)
      
      if (length(lines) <= 6){ #make sure it ran correctly
        #Lines with interesting data
        start_index <- grep("min", lines)
        end_index <- grep("---", lines)
        
        #Extract values
        estimated_Ne <- extract_values(lines[start_index+1])
        computation_time <- extract_values(lines[end_index])[2]
        
        Ne_ONeSAMP <- rbind(Ne_ONeSAMP,
                            c(dataset,pop,subset,"NA",estimated_Ne,computation_time))
      }
      else{
        #add "NA" instead of the values
        Ne_ONeSAMP <- rbind(Ne_ONeSAMP,
                            c(dataset,pop,subset,"Pb_running",rep("NA",7)))
      }
      
}

#cleaning
rm(end_index,start_index,estimated_Ne,text_file,dataset,pop,subset)

# get tab in form
Ne_ONeSAMP <- as.data.frame(Ne_ONeSAMP)
Ne_ONeSAMP$Marker_type <- "snps"
Ne_ONeSAMP <- as.data.frame(Ne_ONeSAMP) %>%
  mutate(across(c("min","max","mean","median","lower95CL","upper95CL","computation_time"), as.numeric))

# check the number of problems while running
cat("Nb of files that did not run correctly:",sum(Ne_ONeSAMP$Pb == "Pb_running"))
```

--> Problem of running have been solved so everyhting was run correctly

Save this table
```{r}
head(Ne_ONeSAMP)

write.csv(Ne_ONeSAMP, file="with_ONeSAMP/summary_results_Ne_withONeSAMP3.csv")
```




