---
title: "Preparation of SNP datasets as Genind datasets"
author: "MGH"
date: "`r Sys.Date()`"
output: html_document
---

The main objective of this code is to obtain usable Genind files from VCF data with the assignation of individuals to populations. This work was mainly a case-by-case work according to the dataset. Indeed, we don't have the same scale of information and neither the same type of information between the species. The approach is described in details in the corresponding part of the Material & Methods, "Population delimitation". 

VCF files used here were obtained after the filtering pipeline (look in 1.SNP_filtering_on_full_files.Rmd for the details). 

We also previously looked at published information from the different studies of the studied datasets (including the original VCF) to create a file containing the list of individuals and their "original" population assignation in the second column, usually corresponding to the sampling location.
We checked below if some sampling locations should be gathered together as one single population. 


# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```

```{r}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
#setwd()
```

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

library(snpR)

library(vcfR)
library(adegenet)

library(graph4lg)
```


# Upload vcf as genind

```{r}
path="vcf_datasets_from_Galaxy/"
filtered_files <- list.files(path, pattern = "\\.vcf$", full.names = TRUE)
basename_files <- sub(path, "", filtered_files)
basename_files <- sub("\\_mdIND_GQ_DP_biallelicSNPs_SNPmd_MAC_mdIND.vcf", "", basename_files)
basename_files
```

So 13 datasets from 11 species. 

```{r}
vcf_2_genind <- function(i){
      
      vcfR_name <-  paste0("Genind_datasets/vcfR/", i,".Rdata")
      gen_name <-  paste0("Genind_datasets/genind_from_vcf_Galaxy/", i,".Rdata")
    
      path2vcf <- paste0(path,i,"_mdIND_GQ_DP_biallelicSNPs_SNPmd_MAC_mdIND.vcf")
        
    if (! file.exists(gen_name)){ #don't run code again if file already exists
      print(i)
      
      #load the vcf
      vcf_file <- read.vcfR(path2vcf)
    
      #conversion into genlight object
      gen_file <- vcfR2genind(vcf_file)
    
      #rename it
      assign(paste0("vcf_",i), vcf_file)
      assign(paste0("gen_",i),gen_file)

      #Save the genind object
      save(list = paste0("vcf_",i),file=vcfR_name)
      save(list = paste0("gen_",i),file=gen_name)
    }
}
```

```{r}
#apply it to all our files (with purr package to avoid loop for)
walk(basename_files,vcf_2_genind)
```



# Correction by removing all loci with Ho > 0.6

Loci with an Ho>0.6 can be the sign of paralog loci. For this reason we will remove them now. 

```{r}
genind_datasets <- list.files("Genind_datasets/genind_from_vcf_Galaxy/", pattern="\\.Rdata$", full.names = TRUE)
genind_datasets <- sub("^Genind_datasets/genind_from_vcf_Galaxy/", "", genind_datasets)
genind_datasets <- sub("\\.Rdata", "", genind_datasets)

for (i in genind_datasets){
  if (file.exists(paste0("Genind_datasets/genind_filtering_Ho_R/",i,".Rdata"))){
    print(paste(i,": already filtered"))
  }else{
    #load the corresponding the genind
    genind_name <- paste0("Genind_datasets/genind_from_vcf_Galaxy/",i,".Rdata")
    loaded <- load(genind_name)
    genind_obj <- get(loaded)
    
    #number of locus at the beginning
    print(nLoc(genind_obj))
    
    #identify the locus to remove
    Ho_locus <- summary(genind_obj)$Hobs
    loci_to_keep <- names(Ho_locus[Ho_locus<0.6])
    print(length(loci_to_keep))
    
    #filter the locus
    genind_obj_filtered <- genind_obj[, loc = loci_to_keep]
    
    print(nLoc(genind_obj_filtered))
  
    #save it
    assign(paste0("gen_",i), genind_obj_filtered)
    save(list=paste0("gen_",i),file=paste0("Genind_datasets/genind_filtering_Ho_R/",i,".Rdata"))
  }
}
```
```{r}
rm(list=ls())
```

# Assign the individuals to their final population 

```{r}
genind_datasets <- list.files("Genind_datasets/genind_filtering_Ho_R/", pattern="\\.Rdata$", full.names = TRUE)
genind_datasets <- sub("^Genind_datasets/genind_filtering_Ho_R/", "", genind_datasets)
genind_datasets <- sub("\\.Rdata", "", genind_datasets)

for (i in genind_datasets){
  #load the corresponding the genind
  genind_name <- paste0("Genind_datasets/genind_filtering_Ho_R/",i,".Rdata")
  loaded <- load(genind_name)
  genind_obj <- get(loaded)
  
  #get the individuals list
  ind_list <- data.frame(Ind = indNames(genind_obj))
  
  #load the population assignation for individuals before the filtering
  pop_assignation <- read.csv(paste0("../Genetic_data_v3/snps_population_assignation/Ind_Pop_list_",i,".txt"))
  pop_assignation <- pop_assignation %>% mutate(Ind = gsub("1_1_", "1_", Ind))

  #A dataset with Ind of the new genind and its pop
  ind_list <- ind_list %>%
      mutate(order = row_number()) %>%  #Index column
      left_join(pop_assignation, by = "Ind") %>%
      arrange(order) %>%  #make sure same order is kept
      select(-order) 
  
    #Put that into the genind and save
    pop(genind_obj) <- as.factor(ind_list$Pop_final)
    print(popNames(genind_obj))
    assign(paste0("gen_",i), genind_obj)
    save(list=paste0("gen_",i),file=paste0("Genind_datasets/genind_with_pop_ass_FINAL/gen_pop_ass_",i,".Rdata"))
}

```

# Last filtering step: Remove populations with n<18

```{r}
genind_datasets <- list.files("Genind_datasets/genind_with_pop_ass_FINAL/", pattern="\\.Rdata$", full.names = TRUE)
genind_datasets <- sub("^Genind_datasets/genind_with_pop_ass_FINAL/gen_pop_ass_", "", genind_datasets)
genind_datasets <- sub("\\.Rdata", "", genind_datasets)

for (i in genind_datasets){
  #load the corresponding the genind
  genind_name <- paste0("Genind_datasets/genind_with_pop_ass_FINAL/gen_pop_ass_",i,".Rdata")
  loaded <- load(genind_name)
  genind_obj <- get(loaded)
  
  #get the name of populations with more than 20 individuals
  pop_counts <- table(genind_obj@pop)
  pops_to_keep <- names(pop_counts[pop_counts >= 18])
  
  #filter the genind
  genind_obj_filtered <- genind_obj[pop(genind_obj) %in% pops_to_keep, ]
  print(table(genind_obj_filtered@pop))
  
  #save it
  assign(paste0("gen_",i), genind_obj_filtered)
  save(list=paste0("gen_",i),file=paste0("Genind_datasets/genind_FINAL/gen_",i,".Rdata"))

}
```



# Transformation from genind to genepop

Convert from genind object in R to genepop files (outside of R)

```{r}
genind_snps_files <- list.files("Genind_datasets/genind_FINAL/", full.names = TRUE)

for (i in genind_snps_files){
  load(i)
  
  genind_data_name <- sub("\\.Rdata$", "", basename(i))
  genind_data <- get(genind_data_name)
  
  
  #cases when only 2 pop (a problem in the iteration of the genind_to_genepop function)
  npop <- nPop(genind_data)
  print(npop)
  
  if (npop == 2){
    #create a false pop
    temp_individuals <- genind_data[1:5]
    pop(temp_individuals) <- factor(rep("Temp_pop", 5))
    
    genind_data <- repool(genind_data, temp_individuals)
    print("REMOVE BY HAND THE ARTIFICIALLY CREATED POP")
  }
    
  #convert
  genepop_name <- sub("^gen_", "", genind_data_name)
  genind_to_genepop(genind_data,output=paste0("Genepop_files/full_snps/",genepop_name,".txt"))
    
}

```


A mistake for Symphonia because some alleles are not biallelic
```{r}
#genind_data_filtered <- genind_data[, loc = genind_data@loc.n.all > 1]
#genind_to_genepop(genind_data_filtered,output=paste0("Genepop_files/",genepop_name,".txt"))
```


First of all, remove the artificially created pop in the files with 2 pop only : Barbastella barbastellus, Cypripedium calceolus and Raja clavata. 

Now we have to reencode the files to make sure alleles are coded into 2-digits. 

Function to use for correction
```{r}
corriger_genotypes <- function(ligne) {
  # Split the name of the individual
  parties <- strsplit(ligne, ", ")[[1]]
  
  if(length(parties) < 2) return(ligne)  # If no genotype, return as it is
  
  nom_individu <- parties[1]
  genotypes_str <- parties[2]
  
  # Split individual genotypes
  genotypes <- strsplit(genotypes_str, " ")[[1]]
  genotypes <- genotypes[genotypes != ""]  # Remove the blank spaces
  
  # Correct each genotype
  genotypes_corriges <- sapply(genotypes, function(gt) {
    if(gt == "000000") {
      return("0000")  # Missing data
    } else if(nchar(gt) == 2) {
      allele1 <- substr(gt, 1, 1)
      allele2 <- substr(gt, 2, 2)
      if(allele1 == "0") allele1 <- "01"
      if(allele1 == "1") allele1 <- "02"
      if(allele2 == "0") allele2 <- "01" 
      if(allele2 == "1") allele2 <- "02"
      
      if(allele1 == "X") allele1 <- "01"
      if(allele1 == "Y") allele1 <- "02"
      if(allele2 == "X") allele2 <- "01" 
      if(allele2 == "Y") allele2 <- "02"
      return(paste0(allele1, allele2))
    } else if(nchar(gt) == 4) {
      return(gt)
    }
  })
  
  # Rebuild the line
  nouvelle_ligne <- paste0(nom_individu, ", ", paste(genotypes_corriges, collapse = " "))
  return(nouvelle_ligne)
}

```


```{r}
# Script pour corriger le format genepop
snps_genepop <- list.files("Genepop_files/full_snps", full.names = TRUE)

for (i in snps_genepop){
  fichier_input <- i  
  fichier_output <- paste0("Genepop_files/full_snps_corrected/",basename(i))

  #Read all the lines
  lignes <- readLines(fichier_input, warn = FALSE)
  
  pop_indices <- grep("^POP", lignes, ignore.case = TRUE)

  #keep the header
  header <- lignes[1:(pop_indices[1]-1)]
  
  # Line by line
  lignes_corrigees <- character(length(lignes))

  for(i in (pop_indices[1]):length(lignes)) {
    ligne <- lignes[i]
  
    # If the line if for an individual, it contains a comma
    if(grepl(",", ligne)) {
    lignes_corrigees[i] <- corriger_genotypes(ligne) #APPLY the correction function
    } else {
    # 
      lignes_corrigees[i] <- trimws(ligne) #If not an individual, do not modify the line
    }
  }

  # Write the corrected file
  writeLines(c(header,lignes_corrigees), fichier_output)

  cat("Corrected and created file:", basename(fichier_output), "\n")
  }

```


# Special case of Ursus arctos

This dataset from Kopatz et al. 2024 was not available at the VCF format but directly as a genind file. 
So, I skipped the filtering part on VCF and will check the missingness of data within the genind file. 

The assignation of individuals to a population was checked and studied using the admixture coefficients of genetic clustering from the original study, combined with geographic data and information from a corresponding SSR dataset (Kopatz et al. 2021). The corresponding codes are not available here but can be shared on request. 

```{r}
raw_data <- read.csv("raw_data/format_other_vcf/Ursus_arctos_Genetic_data_SNPs_Kopatz_2024.csv", sep=";")
info_pop <- read.csv("raw_data/format_other_vcf/Ursus_arctos_Ind_Pop_SNPs_bears_Kopatz_2024.csv", sep=",")

raw_data <- info_pop %>% select(Ind,Population) %>% left_join(raw_data, by="Ind")
table(raw_data$Population)

#construction of adequate dataframe to transform it into a genind object
allele1 <- as.matrix(raw_data[,seq(3,ncol(raw_data)-1,by=2)])
allele2 <- as.matrix(raw_data[,seq(4,ncol(raw_data),by=2)])
gt <- as.data.frame(matrix( paste( allele1, allele2, sep = "/"), ncol = (ncol(raw_data)-2)/2))
colnames(gt)<-colnames(raw_data[,seq(3,ncol(raw_data)-1,by=2)])

#Transform it into a genind object
genind_data <- df2genind(gt, 
                         pop = raw_data$Population, 
                         ind.names = raw_data$Ind,
                         ploidy = 2, 
                         #ncode=3,
                         NA.char = "0",
                         sep="/"
                         )

```

Check quality of data --> missing data and singletons

```{r}
# locus with missing data > 20%? --> Nope
missing_per_locus <- colSums(is.na(tab(genind_data)))/82
which(missing_per_locus > 0.20)

# individuals with missing data > 20%? --> Nope
missing_per_ind <- rowSums(is.na(tab(genind_data)))/93
which(missing_per_ind > 0.20)

```

Almost no missing data!

```{r}
allele_tab <- tab(genind_data)
allele_count <- colSums(allele_tab, na.rm = TRUE)
which(allele_count <= 1)
```

No singletons, so all is good! We can use this dataset directly

```{r}
#rename it
genind_name <-  "gen_Ursus_arctos_Kopatz_2024"
assign(genind_name, genind_data)

#Save the genind object
save(list = genind_name,file=paste0("FINAL_datasets/snps_full/genind_with_pop_ass/",genind_name,".Rdata"))

#genind by pop
  genind_sep <- seppop(genind_data)
  
for (pop in names(genind_sep)){
  genind_obj <- genind_sep[[pop]]
  genind_name <- paste0("gen_Ursus_arctos_Kopatz_2024_",pop)
  assign(genind_name, genind_obj)
  save(list=genind_name,file=paste0("FINAL_datasets/snps_by_pop/genind/",genind_name,".Rdata"))
  
  rm(genind_obj)
}
```

Look at the Fst by pop

```{r}
i="Ursus_arctos_Kopatz_2024"
genl_file <- gi2gl(genind_data)
  
if (file.exists(paste0("raw_data/snps/R_objects/bootstrapped_fst_pop_",i,".Rdata"))){
  load(paste0("raw_data/snps/R_objects/bootstrapped_fst_pop_",i,".Rdata"))
}else{
  #pairwise fst with p-values
  geno_df <- snpR::import.snpR.data(genl_file)
  geno_df <- snpR::calc_pairwise_fst(geno_df,facets="pop",boot=100)

  bootstrapped_fst_pop <- get.snpR.stats(geno_df, "pop", stats ="fst")

  #head(bootstrapped_fst_pop$fst.matrix$pop)
  
  #save this data
  save(bootstrapped_fst_pop,file=paste0("raw_data/snps/R_objects/bootstrapped_fst_pop_",i,".Rdata"))

}

```

```{r}
#turn matrix p-values in a dataframe
Pvalues <- as.data.frame(bootstrapped_fst_pop$fst.matrix$pop$p)
rownames(Pvalues)<- Pvalues[[1]]
Pvalues[[1]] <- NULL
Pvalues <- t(Pvalues)
Pvalues

Pvalues.df <- reshape2::melt(Pvalues, varnames = c('Site1', 'Site2'), value.name="Pvalue", na.rm = TRUE)
head(Pvalues.df)

#turn matrix fst in a dataframe
Fst <- as.data.frame(bootstrapped_fst_pop$fst.matrix$pop$fst)
rownames(Fst)<- Fst[[1]]
Fst[[1]] <- NULL
Fst <- t(Fst)
Fst

Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
head(Fst.df)

#concat
Fst.Pvalues.df <- left_join(Fst.df,Pvalues.df,by=c("Site1","Site2"))
head(Fst.Pvalues.df)
```

```{r}
# Extract middle Fst value for gradient argument
mid = max(Fst.Pvalues.df$Fst) / 2

# Plot heatmap
g <- ggplot(data = Fst.Pvalues.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  geom_text(data = subset(Fst.Pvalues.df, Pvalue > 0.05), aes(label = round(Pvalue, 2)), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 8, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Numbers correspond to p-values for Fst not significantly different from 0")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/snps/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```
