---
title: "Convert microsat files into genind object"
author: "MGH"
date: "2024-10-08"
output:
  html_document:
  toc: true
  toc_float:
  collapsed: false
  smooth_scroll: true
  position: left
  toc_depth: 4
  number_sections: true
  df_print: paged
---

This code aims to convert all the SSR files (in .csv format) into Genind objects. 
These genind object will have the identified individuals and their assigned population. See in the Material & methods of the corresponding publication Harribey et al. 2026, part "Population delimitation" for more details on the choices made for population delimitation. 

To begin with, I gathered all the SSR data in a single folder. 
I modify by hand theses files to have a column with Ind, Population and then columns with SSR data. 
Don't forget to remove special characters, in particular for Population names!!!
All files but two were publicly shared in the ".csv" format. 

# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```

```{r, echo=FALSE}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
#setwd()
```

```{r}
library(dplyr)
library(tidyr)
library(adegenet)
library(poppr)
library(RClone)
library(hierfstat)
```


# Assign individuals to their population

In most cases, the Population of a given individual is known from the original study and can be directly used for the assignation within the Genind object we will create just herafter. 

## Special cases 

### Canis_lupus_Szewczyk_2019

A "population" is called "Putative disperser", gather potential dispersers (see in the paper for more info)
So let's remove it

```{r}
i = "filtered_data/microsat_checked/Canis_lupus_Szewczyk_2019.csv"
raw_data <- read.csv(i, sep=";")
  
#Remove the pop called "putative disperser"
  
raw_data <- raw_data %>% 
              filter(Population != "putative disperser")
print(table(raw_data$Population))
 
 #save this new dataset in a file
j=sub(".*/(.*)\\..*", "\\1", i)
write.table(raw_data, file=paste0("filtered_data/microsat_checked/",j,".csv"), row.names=FALSE, sep=";")
  
```


### Passer domesticus

We have population + sampling year. But as generation time is 1.7 year for that species, we can study one location and one year at a time 
```{r}
i = "filtered_data/microsat_checked/Passer_domesticus_Baalsrud_2013.csv"
raw_data <- read.csv(i, sep=";")

raw_data <- raw_data %>%
  separate(Population, into = c("Location", "Year"), sep = "-", convert = TRUE)

#Count how many populations by year 
raw_data %>%
  distinct(Location, Year) %>%    
  group_by(Year) %>%              
  summarise(n_locations = n())   
```

So I am selecting the year with the more populations = 1994, 2000, 2006

```{r}
raw_data_1994 <- raw_data %>%
                    filter(Year == 1994) %>%
                    unite("Population", Location, Year, sep = "-")

raw_data_2000 <- raw_data %>%
                    filter(Year == 2000) %>%
                    unite("Population", Location, Year, sep = "-")

raw_data_2006 <- raw_data %>%
                    filter(Year == 2006) %>%
                    unite("Population", Location, Year, sep = "-")


#let's save all these files
write.table(raw_data_1994, file="filtered_data/microsat_checked/Passer_domesticus_Baalsrud_2013_pop_1994.csv", row.names=FALSE, sep=";")
write.table(raw_data_2000, file="filtered_data/microsat_checked/Passer_domesticus_Baalsrud_2013_pop_2000.csv", row.names=FALSE, sep=";")
write.table(raw_data_2006, file="filtered_data/microsat_checked/Passer_domesticus_Baalsrud_2013_pop_2006.csv", row.names=FALSE, sep=";")
```


Don't forget to remove the orinal file with locations and years mixed "Passer_domesticus_Baalsrud_2013_pop.csv"

### Mya arenaria

remove individuals with missing data --> in the paper, they remove individuals with more than 4 SSR with missing data

```{r}
#load file
name_new_file="Mya_arenaria_DeNoia_2020"

i=paste0("raw_data/microsat/",name_new_file,".csv")
raw_data <- read.csv(i, sep=";")


#check for individuals with more than 4 loci not genotyped
raw_data <- raw_data[rowSums(is.na(raw_data)) < 8,]
```

Okay so already filtered, nice!


### Coenagrion mercuriale

We will use the results of genetic clustering as a basis for population delimitation (Leveque et al. 2024).

```{r}
#load file
name_new_file="Coenagrion_mercuriale_Leveque_2024"

i=paste0("raw_data/microsat/",name_new_file,".csv")
raw_data <- read.csv("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Raw_genetic_data/new/Coenagrion_mercuriale_Leveque_2024/data_SSRs.csv", sep=";")

#see the pop
table(raw_data$Population)

#column sampling-loc
raw_data$Sampling <- gsub("-.*","",raw_data$Population)
raw_data$Sampling_Year <- gsub(".*-","",raw_data$Population)

#Metapop
name_metapop <- read.delim("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Raw_genetic_data/new/Coenagrion_mercuriale_Leveque_2024/group_pop_genetic_clustering.csv", sep=";")

raw_data <- raw_data %>% left_join(name_metapop, by="Sampling") %>% mutate(Population= case_when(
                        !is.na(Metapop) ~ paste0(Metapop,"-",Sampling_Year),
                                 TRUE ~ paste0(Sampling,"-",Sampling_Year)))

#see the pop
table(raw_data$Population)
length(table(raw_data$Population))

#remove the pop with only one sampling and no cluster assignation
raw_data <- raw_data %>% filter(!Population %in% c("Gox1-2022","S21-2021","S21-2022","Zp0-2022","Zp1-2022","Zp2-2022"))
 #save this new dataset in a file
length(table(raw_data$Population))

write.table(raw_data %>% select(-Sampling,-Sampling_Year,-Metapop), file=paste0("raw_data/microsat/",name_new_file,".csv"), row.names=FALSE, sep=";")
```

### Ursus arctos

```{r}
name_new_file="Ursus_arctos_Kopatz_2021"

i=paste0("raw_data/microsat/",name_new_file,".csv")
raw_data <- read.csv(i, sep=";")

table(raw_data$Population)
```

Population assignation checked in details, but not presented here. The assignation of individuals to a population was checked and studied using the admixture coefficients of genetic clustering from the original study, combined with geographic data and information from a corresponding SNP dataset (Kopatz et al. 2024). The corresponding codes are not available here but can be shared on request. 


There are individuals with only missing data so let's remove them now to don't have trouble later

```{r}
raw_data <- raw_data %>% drop_na(all_of(names(raw_data)[3:26]))

write.table(raw_data, file="filtered_data/microsat_checked/Ursus_arctos_Kopatz_2021.csv", row.names=FALSE, sep=";")
```



# Get a Genind object

This format of file is useful to treat with DNA dataset (and will be a common format between snp and microsat data)
I am going to conserve this genind object into an .Rdata file

```{r}
files_list <- list.files("filtered_data/microsat_checked/", full.names = TRUE)

#files_list <- files_list[files_list != "filtered_data/microsat_checked/Canis_lupus_Szewczyk_2019_pop.csv"]

for (i in files_list){
  
  #load the dataset
  print(i)
  raw_data <- read.csv(i, sep=";") #check to be sure
  
  

  #construction of adequate dataframe to transform it into a genind object
  allele1 <- as.matrix(raw_data[,seq(3,ncol(raw_data)-1,by=2)])
  allele2 <- as.matrix(raw_data[,seq(4,ncol(raw_data),by=2)])
  gt <- as.data.frame(matrix( paste( allele1, allele2, sep = "/"), ncol = (ncol(raw_data)-2)/2))
  colnames(gt)<-colnames(raw_data[,seq(3,ncol(raw_data)-1,by=2)])

  #Transform it into a genind object
  genind_data <- df2genind(gt, 
                         pop = raw_data$Population, 
                         ind.names = raw_data$Ind,
                         ploidy = 2, 
                         #ncode=3,
                         NA.char = "0",
                         sep="/"
                         )

  #rename it
  genind_name <-  paste0("genind_", sub(".*/(.*)\\..*", "\\1", i))
  assign(genind_name, genind_data)

  #Save the genind object
  save(list = genind_name,file=paste0("FINAL_datasets/microsat/",genind_name,".Rdata"))
  
  #cleaning
  rm(allele1,allele2,gt,genind_data)

}
 
```


## Other format of data

Two datasets were already provided as Genepop files: Alytes obstetricans, Barratt et al. 2024 and Graellsia isabellae, Gonzales-Castellano et al. 2023. The file could directly be loaded as a Genind object with the function "read_genepop" of adegenet. 


```{r}
i="Graellsia_isabellae_GonzalesCastellano_2023"
genind_graell <- read.genepop("raw_data/microsat/Graellsia_isabellae_GonzalesCastellano_2023.gen")

#check
nPop(genind_graell)
nInd(genind_graell)
nLoc(genind_graell)
```

```{r}
i="Alytes_obstetricans_Barratt_2024"
genind_alyt <- read.genepop("raw_data/microsat/Alytes_obstetricans_Barratt_2024.gen",ncode=3)

#check
nPop(genind_alyt)
nInd(genind_alyt)
nLoc(genind_alyt)

pop_sizes <- table(pop(genind_alyt))

```


# Special case of Cypripedium calceolus: clone correction

See in Gargiulo et al. (2023) --> the raw SSRs dataset need a correction for clones
MLG or MLL clone correction is possible + HW equilibirum correction
Using code from https://github.com/anchampion/Ne_estimation_EcM_fungi

```{r}
rm(list=ls())
```

```{r}
load("FINAL_datasets/microsat/genind_Cypripedium_calceolus_Gargiulo_2023.Rdata")
genind_Cypripedium_calceolus_Gargiulo_2023
```

Correction for clone with poppr package

```{r}
data.df <- genind2df(genind_Cypripedium_calceolus_Gargiulo_2023)
popvec <- data.df[,1] # pop info (if present)
dataset <- data.df[,2:ncol(data.df)]# dataset
dataset <- as.data.frame(apply(dataset, 2, function(x) replace(x, is.na(x), "000000"))) # replace NAs by 000000
data2 <- convert_GC(dataset, 3) # we use "3" because this is the length of the allele
row.names(data2) <- seq_len(nrow(data2)) # make rownames consecutive integers starting from 1 to avoid bugs
head(data2)
```

```{r}
# List unique alleles per locus
list_all_tab(data2)
```

Determining MLGs. 

```{r}
MLGlist <- MLG_list(data2)
# Allelic frequencies :
freq_RR(data2)
```

```{r}
#genetic distances computation, distance on allele differences:
respop <- genet_dist(data2) # max 10 differences between genotypes
```


```{r}
# Extract one individual from each MLG 
new_ind <- sapply(MLGlist, `[[`, 1)
new_ind <- as.numeric(new_ind);new_ind
```

```{r}
rownames(genind_Cypripedium_calceolus_Gargiulo_2023@tab)<-c(1:1123) # if ids are not set to 1,2,3,...n
selected_indices <- match(new_ind, rownames(genind_Cypripedium_calceolus_Gargiulo_2023@tab));selected_indices
```

```{r}
# Create a new genind object without the clonal individuals
data_noclones <- genind_Cypripedium_calceolus_Gargiulo_2023[selected_indices, ]
# Check if the genind object is correct
data_noclones
```

Let's save it as our new genind object

```{r}
genind_Cypripedium_calceolus_Gargiulo_2023 <- data_noclones
genind_Cypripedium_calceolus_Gargiulo_2023

save(genind_Cypripedium_calceolus_Gargiulo_2023, file=("FINAL_datasets/microsat/genind_Cypripedium_calceolus_Gargiulo_2023.Rdata"))
```



# Check pairwise differentiation between sampling location

If there is no significant differences and the locations are geographically close, we can consider gathering the sampling locations together and consider them as a single population. 

### Mya arenaria

```{r}
i="Mya_arenaria_DeNoia_2020"
load("FINAL_datasets/microsat/genind_Mya_arenaria_DeNoia_2020.Rdata")

genind_mya <- genind_Mya_arenaria_DeNoia_2020

#check
nPop(genind_mya)
nInd(genind_mya)
nLoc(genind_mya)
```
Genetic differentiation between pop
```{r}
pairw_fst_mya <- genet.dist(genind_mya, diploid=TRUE, method="WC84")
pairw_fst_mya

#turn matrix fst in a dataframe
Fst <- as.matrix(pairw_fst_mya)
Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
Fst.df <- Fst.df %>% distinct()
head(Fst.df)

#Check significance
bornes_fst <- boot.ppfst(genind_mya, nboot = 1000)
non_significant_fst <- (bornes_fst$ll < 0) & (bornes_fst$ul > 0) # lower limit < 0 and upper limit > 0 = contains 0
pop_non_sign <- which(non_significant_fst == TRUE, arr.ind = TRUE)
pop_non_sign <- data.frame(
  Site1 = rownames(non_significant_fst)[pop_non_sign[,1]],
  Site2 = colnames(non_significant_fst)[pop_non_sign[,2]]
)
pop_non_sign$Sign <- "NS"
pop_non_sign <- pop_non_sign %>%
  bind_rows(pop_non_sign %>%
              rename(Site1 = Site2, Site2 = Site1))
pop_non_sign
#empty so no pop non sign


g <- ggplot(data = Fst.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  #geom_text(data = subset(Fst.Pvalues.df, Sign == "NS"), aes(label = Sign), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 8, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Pairwise Fst significantly diff from 0 = NS")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/microsat/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```

### Rana arvalis

Here sampling locations correspond to patches, and can be very close. Note that the sampling area here is very restricted to one sample zone in the area. 

```{r}
i="Rana_arvalis_Cox_2024"
load("FINAL_datasets/microsat/genind_Rana_arvalis_Cox_2024.Rdata")

genind_rana <- genind_Rana_arvalis_Cox_2024

#check
nPop(genind_rana)
nInd(genind_rana)
nLoc(genind_rana)
```

Genetic differentiation between pop
```{r}
pairw_fst_rana <- genet.dist(genind_rana, diploid=TRUE, method="WC84")
pairw_fst_rana

#turn matrix fst in a dataframe
Fst <- as.matrix(pairw_fst_rana)
Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
Fst.df <- Fst.df %>% distinct()
head(Fst.df)

#Check significance
bornes_fst <- boot.ppfst(genind_rana, nboot = 1000)
non_significant_fst <- (bornes_fst$ll < 0) & (bornes_fst$ul > 0) # lower limit < 0 and upper limit > 0 = contains 0
pop_non_sign <- which(non_significant_fst == TRUE, arr.ind = TRUE)
pop_non_sign <- data.frame(
  Site1 = rownames(non_significant_fst)[pop_non_sign[,1]],
  Site2 = colnames(non_significant_fst)[pop_non_sign[,2]]
)
pop_non_sign$Sign <- "NS"
pop_non_sign <- pop_non_sign %>%
  bind_rows(pop_non_sign %>%
              rename(Site1 = Site2, Site2 = Site1))
pop_non_sign

#join the two tabs in one
Fst.Pvalues.df <- left_join(Fst.df,pop_non_sign,by=c("Site1","Site2"))
head(Fst.Pvalues.df)


g <- ggplot(data = Fst.Pvalues.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  geom_text(data = subset(Fst.Pvalues.df, Sign == "NS"), aes(label = Sign), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 8, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Pairwise Fst significantly diff from 0 = NS")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/microsat/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```

So we will gather some of the patches together as single population. 

```{r}
ind_pop_tab <- data.frame(Ind=indNames(genind_data),
                          Pop_or=pop(genind_data))

ind_pop_tab <- ind_pop_tab %>%
                dplyr::mutate(Pop_final=case_when(
                  Pop_or %in% c("patch_22A","patch_22B","patch_22C","patch_25") ~ "patch_22_25" , ############# TO BE MODIFIED #############
                   Pop_or %in% c("patch_3","patch_4","patch_5","patch_6","patch_7") ~ "patch_3_4_5_6_7" ,
                  Pop_or %in% c("patch_10","patch_11") ~ "patch_10_11" ,
                  Pop_or %in% c("patch_13","patch_14") ~ "patch_13_14" ,
                  Pop_or %in% c("patch_16","patch_17") ~ "patch_16_17" ,
                  TRUE ~ Pop_or
                ))


table(ind_pop_tab$Pop_final)


#assign this pop names to genind object
pop(genind_data) <- ind_pop_tab$Pop_final
```

```{r}
pop_counts <- table(genind_data@pop)
  pops_to_keep <- names(pop_counts[pop_counts >= 20])
  
  #filter the genind
  genind_data_filtered <- genind_data[pop(genind_data) %in% pops_to_keep, ]
  print(table(genind_data_filtered@pop))
  
  #save it
  assign(paste0("gen_",i), genind_data_filtered)
  save(list=paste0("gen_",i),file=paste0("gen_",i,".Rdata"))
  write.csv(ind_pop_tab, file="Ind_pop_assignation_Rana_arvalis.csv")
```

### Gasterosteus aculeatus

```{r}
i="Gasterosteus_aculeatus_Raeymaekers_2008"
load("FINAL_datasets/microsat/genind_Gasterosteus_aculeatus_Raeymaekers_2008.Rdata")

genind_gast <- genind_Gasterosteus_aculeatus_Raeymaekers_2008

#check
nPop(genind_gast)
nInd(genind_gast)
nLoc(genind_gast)
```
Genetic differentiation between pop
```{r}
pairw_fst_gast <- genet.dist(genind_gast, diploid=TRUE, method="WC84")
pairw_fst_gast

#turn matrix fst in a dataframe
Fst <- as.matrix(pairw_fst_gast)
Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
Fst.df <- Fst.df %>% distinct()
head(Fst.df)

#Check significance
bornes_fst <- boot.ppfst(genind_gast, nboot = 1000)
non_significant_fst <- (bornes_fst$ll < 0) & (bornes_fst$ul > 0) # lower limit < 0 and upper limit > 0 = contains 0
pop_non_sign <- which(non_significant_fst == TRUE, arr.ind = TRUE)
pop_non_sign <- data.frame(
  Site1 = rownames(non_significant_fst)[pop_non_sign[,1]],
  Site2 = colnames(non_significant_fst)[pop_non_sign[,2]]
)
pop_non_sign
#empty so no pop non sign


g <- ggplot(data = Fst.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  #geom_text(data = subset(Fst.Pvalues.df, Sign == "NS"), aes(label = Sign), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 8, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Pairwise Fst significantly diff from 0 = NS")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/microsat/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```

All populations are significantly genetically different. 


### Alytes obstetricans

```{r}
i="Alytes_obstetricans_Barratt_2024"
genind_alyt <- read.genepop("raw_data/microsat/Alytes_obstetricans_Barratt_2024.gen",ncode=3)

#check
nPop(genind_alyt)
nInd(genind_alyt)
nLoc(genind_alyt)

pop_sizes <- table(pop(genind_alyt))
valid_pops <- names(pop_sizes[pop_sizes >= 20])

```

```{r}
pairw_fst_alyt_loc <- genet.dist(genind_alyt, diploid=TRUE, method="WC84")
pairw_fst_alyt_loc


#turn matrix fst in a dataframe
Fst <- as.matrix(pairw_fst_alyt_loc)
Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
Fst.df <- Fst.df %>% distinct()
head(Fst.df)

#Check significance
bornes_fst <- boot.ppfst(genind_alyt, nboot = 1000)
non_significant_fst <- (bornes_fst$ll < 0) & (bornes_fst$ul > 0) # lower limit < 0 and upper limit > 0 = contains 0
pop_non_sign <- which(non_significant_fst == TRUE, arr.ind = TRUE)
pop_non_sign <- data.frame(
  Site1 = rownames(non_significant_fst)[pop_non_sign[,1]],
  Site2 = colnames(non_significant_fst)[pop_non_sign[,2]]
)
pop_non_sign$Sign <- "NS"
pop_non_sign <- pop_non_sign %>%
  bind_rows(pop_non_sign %>%
              rename(Site1 = Site2, Site2 = Site1))
pop_non_sign

#join the two tabs in one
Fst.Pvalues.df <- left_join(Fst.df,pop_non_sign,by=c("Site1","Site2"))
head(Fst.Pvalues.df)

g <- ggplot(data = Fst.Pvalues.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  geom_text(data = subset(Fst.Pvalues.df, Sign == "NS"), aes(label = Sign), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 6, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Pairwise Fst significantly diff from 0 = NS")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/microsat/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```

Don't find exactly the same association that we could expect with genetic clustering

Okay, I will gather some populations when there is no genetic diff + very close geographically + belong to the same genetic cluster

```{r}
ind_pop_tab <- data.frame(Ind=indNames(genind_alyt),
                          Pop_or=pop(genind_alyt))

ind_pop_tab <- ind_pop_tab %>%
                dplyr::mutate(Pop_final=case_when(
                  Pop_or %in% c("pop7","pop11","pop12","pop21","pop23") ~ "MetapopA" , ############# TO BE MODIFIED #############
                   Pop_or %in% c("pop26","pop27") ~ "MetapopB" ,
                  Pop_or %in% c("pop8","pop9","pop33") ~ "MetapopC" ,
                  TRUE ~ Pop_or
                ))


table(ind_pop_tab$Pop_final)


#assign this pop names to genind object
pop(genind_alyt) <- ind_pop_tab$Pop_final
```


Otherwise, simply keep only the locations with more than 20 ind

```{r}
pop_sizes <- table(pop(genind_alyt))
valid_pops <- names(pop_sizes[pop_sizes >= 20])
valid_pops

#filtering that pops
genind_alyt <- genind_alyt[pop(genind_alyt) %in% valid_pops]
nPop(genind_alyt)
nInd(genind_alyt)
nLoc(genind_alyt)

#rename it
genind_name <-  paste0("genind_",i)
assign(genind_name, genind_alyt)

#Save the genind object
save(list = genind_name,file=paste0("FINAL_datasets/microsat/",genind_name,".Rdata"))
```

### Graellsia isabellae 

```{r}
i="Graellsia_isabellae_GonzalesCastellano_2023"
genind_graell <- read.genepop("raw_data/microsat/Graellsia_isabellae_GonzalesCastellano_2023.gen")

#check
nPop(genind_graell)
nInd(genind_graell)
nLoc(genind_graell)
```

Genetic differentiation between population 
```{r}
pairw_fst_graell <- genet.dist(genind_graell, diploid=TRUE, method="WC84")
pairw_fst_graell

#turn matrix fst in a dataframe
Fst <- as.matrix(pairw_fst_graell)
Fst.df <- reshape2::melt(Fst, varnames = c('Site1', 'Site2'), value.name="Fst", na.rm = TRUE)
Fst.df <- Fst.df %>% distinct()
head(Fst.df)

#Check significance
bornes_fst <- boot.ppfst(genind_graell, nboot = 1000)
non_significant_fst <- (bornes_fst$ll < 0) & (bornes_fst$ul > 0) # lower limit < 0 and upper limit > 0 = contains 0
pop_non_sign <- which(non_significant_fst == TRUE, arr.ind = TRUE)
pop_non_sign <- data.frame(
  Site1 = rownames(non_significant_fst)[pop_non_sign[,1]],
  Site2 = colnames(non_significant_fst)[pop_non_sign[,2]]
)
pop_non_sign$Sign <- "NS"
pop_non_sign <- pop_non_sign %>%
  bind_rows(pop_non_sign %>%
              rename(Site1 = Site2, Site2 = Site1))
pop_non_sign

#join the two tabs in one
Fst.Pvalues.df <- left_join(Fst.df,pop_non_sign,by=c("Site1","Site2"))
head(Fst.Pvalues.df)


g <- ggplot(data = Fst.Pvalues.df, aes(x = Site1, y = Site2, fill = Fst))+
  geom_tile(colour = "black")+
  geom_text(data = subset(Fst.Pvalues.df, Sign == "NS"), aes(label = Sign), color = "black", size = 3)+
  scale_fill_gradient2(low = "#4B0055FF", mid = "#009796FF", high = "#FDE333FF")+
  scale_x_discrete(expand = c(0,0), )+
  scale_y_discrete(expand = c(0,0), position = "right")+
  theme(axis.text = element_text(colour = "black", size = 8, angle=45, vjust=1, hjust=1),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_blank(),
        legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
        )+
  ggtitle(bquote(
    atop(.(i), 
    atop("Pairwise Fst",italic("Pairwise Fst significantly diff from 0 = NS")))
    ))
print(g)
ggsave(g,file=paste0("raw_data/microsat/Population_assignation/Matrix_pairwise_Fst_",i,".pdf"))
```

Okay, I will gather some populations: BUJ-ORD-COT and PLA-ESC-TEL
Because no genetic diff + very close geographically

```{r}
ind_pop_tab <- data.frame(Ind=indNames(genind_graell),
                          Pop_or=pop(genind_graell))

ind_pop_tab <- ind_pop_tab %>%
                dplyr::mutate(Pop_final=case_when(
                  Pop_or %in% c("BUJ","ORD","COT") ~ "BUJ-ORD-COT" , ############# TO BE MODIFIED #############
                   Pop_or %in% c("PLA","ESC","TEL") ~ "PLA-ESC-TEL" ,
                  TRUE ~ Pop_or
                ))


table(ind_pop_tab$Pop_final)


#assign this pop names to genind object
pop(genind_graell) <- ind_pop_tab$Pop_final
```


Otherwise, simply keep only the locations with more than 20 ind

```{r}
pop_sizes <- table(pop(genind_graell))
valid_pops <- names(pop_sizes[pop_sizes >= 20])
valid_pops

#filtering that pops
genind_graell <- genind_graell[pop(genind_graell) %in% valid_pops]
nPop(genind_graell)
nInd(genind_graell)
nLoc(genind_graell)

#rename it
genind_name <-  paste0("genind_",i)
assign(genind_name, genind_graell)

#Save the genind object
save(list = genind_name,file=paste0("FINAL_datasets/microsat/",genind_name,".Rdata"))
```


# Filter population with n<20

```{r}

               
               
files_list <- paste0("raw_data/microsat/",new_files,".csv")
files_list
```


```{r}
genind_datasets <- list.files("FINAL_datasets/microsat/", pattern="\\.Rdata$", full.names = TRUE)
genind_datasets <- sub("^FINAL_datasets/microsat/", "", genind_datasets)
genind_datasets <- sub("\\.Rdata", "", genind_datasets)

#new files
#new_files <- c(#"Antirrhinum_charidemi_HeuertzForrest_2024",
               #"Rana_arvalis_Cox_2024",
               #"Mya_arenaria_DeNoia_2020")
                #"Coenagrion_mercuriale_Leveque_2024")
                #"Gasterosteus_aculeatus_Raeymaekers_2008")

for (i in genind_datasets){
  #load the corresponding the genind
  genind_name <- paste0("FINAL_datasets/microsat/",i,".Rdata")
  loaded <- load(genind_name)
  genind_obj <- get(loaded)
  
  #get the name of populations with more than 20 individuals
  pop_counts <- table(genind_obj@pop)
  pops_to_keep <- names(pop_counts[pop_counts >= 20])
  
  #filter the genind
  genind_obj_filtered <- genind_obj[pop(genind_obj) %in% pops_to_keep, ]
  print(table(genind_obj_filtered@pop))
  
  #save it
  assign(paste0("gen_",i), genind_obj_filtered)
  save(list=paste0("gen_",i),file=paste0("FINAL_datasets/microsat_final/",i,".Rdata"))

}
```

