---
title: "Summary statistics for SNPs datasets (Genetic data v4)"
author: "MGH"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 4
    number_sections: true
    df_print: paged
---

Now, we are going to calculate the summary statistics for SNP datasets. We will include a process of parallelization within R, in order to speed up the calculations. 
Calculation are splitted into two parts:
- Estimates within a population: He, Ho, Ar, Fis
- Pairwise estimates and then averaging for a given population within a dataset: Fst, Gst', Jost's D

# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```

```{r}
rm(list=ls())
setwd("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v4")
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)

library(parallel)

library(adegenet)
library(factoextra)
library(pegas)
library(poppr)
library(hierfstat)
library(mmod)
```

Find the folder with the list of genind files to use for the calculation. 

```{r}
genind_snps_files <- list.files("Genind_datasets/genind_FINAL", pattern = "\\.Rdata$", full.names = TRUE)
marker_type="snps"

```

# Summary stats by pop


```{r}
compute_diversity_stats <- function(filepath) {
  tryCatch({
    
    load(filepath)
    genind_data_name <- sub("\\.Rdata$", "", basename(filepath))
    genind_data <- get(genind_data_name)
    
    #to follow what's going on
    marker_file <- paste0("Summary_statistics/state_parallel_sum_div/progress_", basename(filepath), ".txt")
    writeLines(paste("Started at", Sys.time()), con = marker_file)
    
    pops_list <- seppop(genind_data, drop = TRUE)
    
    n_ind_per_pop <- sapply(pops_list, nInd)
    
    #locus per pop
    n_loci_per_pop <- sapply(pops_list, nLoc)
    max_loci <- max(n_loci_per_pop)
    
    # Heterozygosity
    Hobs <- t(sapply(pops_list, function(ls) {
                                  hobs_vec <- summary(ls)$Hobs
                                  #to ensure we calculate it for all variants across all populations
                                  length(hobs_vec) <- max_loci
                                  return(hobs_vec)
              }))
    Hexp <- t(sapply(pops_list, function(ls) {
                                  hexp_vec <- summary(ls)$Hexp
                                  #to ensure we calculate it for all variants across all populations
                                  length(hexp_vec) <- max_loci 
                                  return(hexp_vec)
                                  }))
    
    Hobs.pop <- apply(Hobs, 1, mean, na.rm = TRUE)
    Hexp.pop <- apply(Hexp, 1, mean, na.rm = TRUE)
    
    names(Hobs.pop) <- levels(genind_data@pop)
    names(Hexp.pop) <- levels(genind_data@pop)
    
    #pi Milesi
    sum.Hexp.pop <- apply(Hexp, 1, sum, na.rm = TRUE)
    n_all <- table(genind_data@pop)*2 #number of allele copies
    L <- nLoc(genind_data)
    pi_milesi <- (n_all/(L*(n_all-1)))*sum.Hexp.pop
    
    cat(paste("Ho, He, Pi ok", Sys.time()), file=marker_file, append=TRUE)
    
    # Allelic Richness
    Richness <- allelic.richness(genind_data, diploid = TRUE)
    Richness_mean <- colMeans(as.matrix(Richness$Ar), na.rm = TRUE)
    
    cat(paste("Ar ok", Sys.time()), file=marker_file, append=TRUE)
    
    # Fis
    Fis <- t(sapply(seppop(genind_data), function(ls) basic.stats(ls)$perloc$Fis))
    Fis.pop <- rowMeans(as.matrix(Fis), na.rm = TRUE)
    
    cat(paste("Fis ok", Sys.time()), file=marker_file, append=TRUE)
      
    # DEBUG: log the structure before data.frame creation
    cat(sprintf("Hobs.pop length: %d\n", length(Hobs.pop)), file = marker_file, append = TRUE)
    cat(sprintf("Hexp.pop length: %d\n", length(Hexp.pop)), file = marker_file, append = TRUE)
    cat(sprintf("pi_milesi length: %d\n", length(pi_milesi)), file = marker_file, append = TRUE)
    cat(sprintf("Richness_mean length: %d\n", length(Richness_mean)), file = marker_file, append = TRUE)
    cat(sprintf("Fis.pop length: %d\n", length(Fis.pop)), file = marker_file, append = TRUE)
    cat(sprintf("n.by.pop length: %d\n", length(Sum$n.by.pop)), file = marker_file, append = TRUE)
    
    
    return(list(
      "Pop" = names(Hobs.pop),
      "n_ind" = as.vector(n_ind_per_pop),
      "n_loci" = as.vector(n_loci_per_pop),
      "Hobs" = as.vector(Hobs.pop),
      "Hexp" = as.vector(Hexp.pop),
      "Pi_Milesi"= as.vector(pi_milesi),
      "Ar" = as.vector(Richness_mean),
      "Fis" = as.vector(Fis.pop),
      "Dataset" = genind_data_name,
      "Marker_type" = marker_type,
      "status" = "success"
    ))
    
    
  }, error = function(e) {
    cat(paste("Failed to run", Sys.time()), file = marker_file, append = TRUE)
    return(list(
      "data" = NULL,
      "status" = "failed",
      "error" = e$message,
      "file" = basename(filepath)))
  })
  
  #info about the time
  cat(paste("Finished at", Sys.time()), file=marker_file, append=TRUE)
}

# for the parallelisation
n_cores <- detectCores() - 1
cl <- makeCluster(n_cores, type = "PSOCK")
# Load packages on workers
clusterEvalQ(cl, {
  library(adegenet)
  library(hierfstat)
})

clusterExport(cl, c("compute_diversity_stats", "marker_type"), envir = environment())

# run in parallel
Summary_div_stats <- parLapplyLB(cl, genind_snps_files, compute_diversity_stats)
names(Summary_div_stats) <- sub("\\.Rdata$", "", basename(genind_snps_files))

```

```{r}
#clean the cluster
stopCluster(cl)
gc() 
```



We need to check if there are some problems for some dataset
```{r}
failed <- sapply(Summary_div_stats, function(x) {
  is.null(x) || x$status != "success"
})
Summary_div_stats[failed]
```

Save the file
```{r}
save(Summary_div_stats, file="Summary_statistics/snps_summary_div_stats.Rdata")
```


Now store every results in a single tab

```{r}
df_sum_div_stats <- do.call(rbind, lapply(Summary_div_stats, function(x) {
    do.call(cbind, x)
  }))
df_sum_div_stats <- as.data.frame(df_sum_div_stats)

#save the table
write.csv(df_sum_div_stats, file="Summary_statistics/snps_summary_div_stats_corrected.csv")
```


# Pairwise values for differentiation stats

## Get the pairwise values

```{r}
#Function to compute pairwise
compute_pairwise <- function(filepath) {
  tryCatch({
    #get the the genind file
    load(filepath)
    genind_data_name <- sub("\\.Rdata$", "", basename(filepath))
    dataset_all_pop <- get(genind_data_name)
    
    #to follow what's going on
    marker_file <- paste0("Summary_statistics/progress_", basename(filepath), ".txt")
    writeLines(paste("Started at", Sys.time()), marker_file)
    
    #estimate the pairwises
    fst <- genet.dist(dataset_all_pop, diploid = TRUE, method = "WC84")
    writeLines(paste("fst ok", Sys.time()), marker_file)
    
    gst_pr_nei <- pairwise_Gst_Nei(dataset_all_pop)
    writeLines(paste("gst ok", Sys.time()), marker_file)
    
    jost_D <- pairwise_D(dataset_all_pop)
    writeLines(paste("jost ok", Sys.time()), marker_file)
    
    #load them in a list
    return(list(
      "fst" = fst,
      "gst_pr_nei" = gst_pr_nei,
      "jost_D" = jost_D,
      "status" = "success"
    ))
  }, error = function(e) {
    # Return error info instead of crashing
    return(list(
      "error" = as.character(e),
      "status" = "failed"
    ))
  })
  
  #info about the time
  elapsed <- difftime(Sys.time(), start_time, units = "secs")
  writeLines(paste("Finished at", Sys.time()), marker_file)
  writeLines(paste("In total, last for ", elapsed), marker_file)
  
}

# ask parallel execution
n_cores <- detectCores() - 2
cl <- makeCluster(n_cores, type = "PSOCK")
clusterEvalQ(cl, { #load the packages you'll need
  library(adegenet)
  library(mmod)
  library(hierfstat)
})
clusterExport(cl, "compute_pairwise", envir = environment())

#apply it on the genind files
snps_Pairwise_values_by_dataset <- parLapplyLB(cl, genind_snps_files, compute_pairwise)
names(snps_Pairwise_values_by_dataset) <- sub("\\.Rdata$", "", basename(genind_snps_files))

```

```{r}
# Check which ones failed
failed <- sapply(snps_Pairwise_values_by_dataset, function(x) x$status == "failed")
if (any(failed)) {
  message("Failed files:")
  print(genind_snps_files[failed])
  print(sapply(results[failed], function(x) x$error))
}

#clean the cluster
stopCluster(cl)
gc() 
```


Save all
```{r}
save(snps_Pairwise_values_by_dataset,file="Summary_statistics/snps_Pairwise_values_by_dataset.Rdata")
```


## Average pairwise values for a given pop


```{r}

mean_Pairwise_values_by_pop_snps <- matrix(ncol=4, nrow=0)

for (i in names(snps_Pairwise_values_by_dataset)){
  stats_diff <- c("fst","gst_pr_nei","jost_D")
  for (measure in stats_diff){
    
    labels <- attr(snps_Pairwise_values_by_dataset[[i]][[measure]], "Labels")
    values <- as.matrix(snps_Pairwise_values_by_dataset[[i]][[measure]])
    
    for (pop in labels){
      all_pairwise_1pop <- values[pop, ]
      all_pairwise_1pop <- all_pairwise_1pop[names(all_pairwise_1pop) != pop] #to remove pop itself
      
      mean_Pairwise_values_by_pop_snps <- rbind(mean_Pairwise_values_by_pop_snps,
                                           c(Dataset = i,
                                              Diff_stat = measure, 
                                              Pop = pop,
                                              mean_pairwise = mean(all_pairwise_1pop)))
    }
    
  }
}
rm(i,stats_diff,measure,labels,values,pop,all_pairwise_1pop)

mean_Pairwise_values_by_pop_snps <- as.data.frame(mean_Pairwise_values_by_pop_snps)
mean_Pairwise_values_by_pop_snps$mean_pairwise <- as.numeric(mean_Pairwise_values_by_pop_snps$mean_pairwise)

mean_Pairwise_values_by_pop_snps <- mean_Pairwise_values_by_pop_snps %>%
                                                pivot_wider(names_from = Diff_stat, values_from = mean_pairwise)

# there is a just a problem in the name of the pop for Symphonia, let's rename them
mean_Pairwise_values_by_pop_snps <- mean_Pairwise_values_by_pop_snps %>% mutate(Pop=case_when(
                                                          Pop == "gen_Symphonia_globulifera_Schmitt2021_Gargiulo2024_Species1" ~ "Species1",
                                                          Pop == "gen_Symphonia_globulifera_Schmitt2021_Gargiulo2024_Species2" ~ "Species2",
                                                          Pop == "gen_Symphonia_globulifera_Schmitt2021_Gargiulo2024_Species3" ~ "Species3",
                                                          TRUE ~ Pop
))


#check that it's all good
head(mean_Pairwise_values_by_pop_snps)
tail(mean_Pairwise_values_by_pop_snps)
```

```{r}
#saving
write.table(mean_Pairwise_values_by_pop_snps, file="Summary_statistics/snps_mean_Pairwise_values_by_pop.csv", sep=";")
```

# Save the data in a single tab

Gather the tabs in a single one

Add this to the summary diversity statistics 

```{r}
Summary_div_stats <- read.csv("Summary_statistics/snps_summary_div_stats.csv")
mean_Pairwise_values_by_pop_snps <- read.csv("Summary_statistics/snps_mean_Pairwise_values_by_pop.csv", sep=";")

Summary_div_stats_snps <- Summary_div_stats %>% left_join(mean_Pairwise_values_by_pop_snps %>% select(-Dataset), by="Pop")

# CHECK IT'S ALL GOOD BEFORE SAVING
write.csv(Summary_div_stats_snps, file="Summary_statistics/snps_summary_div_diff_stats_by_pop.csv", row.names = FALSE)
```


