---
title: "Summary statistics fro SSR datasets"
author: "MGH"
date: "2024-10-25"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
      position: left
    toc_depth: 4
    number_sections: true
    df_print: paged
---


The objective here is to estimate genetic summary statistics for our different datasets in the SSR format.
All datasets have been converted into a genind format.

Calculation are splitted into two parts:
- Estimates within a population: He, Ho, Ar, Fis
- Pairwise estimates and then averaging for a given population within a dataset: Fst, Gst', Jost's D


# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```


```{r}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
#setwd()
```

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(purrr)
library(corrplot)
library(GGally)

#specific to pop genetics
library(adegenet)
library(factoextra)
library(pegas)
library(poppr)
library(hierfstat)
library(mmod)
```


Set up
```{r}
microsat_files_list <- list.files("FINAL_datasets/microsat/", pattern = "\\.Rdata$", full.names = TRUE)
marker_type="microsat"


# A problem with this dataset because one of the pop has one microsat less
microsat_files_list <- microsat_files_list[microsat_files_list != "FINAL_datasets/microsat/genind_Antirrhinum_charidemi_HeuertzForrest_2024.Rdata"]

```


# Summary stats by pop

```{r}
Summary_div_stats <- as.data.frame(matrix(ncol=8,nrow=0))
colnames(Summary_div_stats) <- c("Pop","n","Hobs","Hexp","Ar","Fis","Dataset","Marker_type")

for (i in microsat_files_list){
  load(i)
  genind_data_name <- sub("\\.Rdata$", "", basename(i))
  genind_data <- get(genind_data_name)
  
  Sum <- summary(genind_data)

  #Heterozygosity
  Hobs <- t(sapply(seppop(genind_data,drop=TRUE), function(ls) summary(ls)$Hobs))
  Hexp <- t(sapply(seppop(genind_data,drop=TRUE), function(ls) summary(ls)$Hexp))
  
  Hobs.pop <- apply(Hobs, 1, mean, na.rm=TRUE)
  Hexp.pop <- apply(Hexp, 1, mean, na.rm=TRUE)
  #adding population names
  names(Hobs.pop) <- levels(genind_data@pop)
  names(Hexp.pop) <- levels(genind_data@pop)

  #Allelic Richness
  Richness <- hierfstat::allelic.richness(genind_data, diploid=TRUE)

  Richness_mean <- colMeans(as.matrix(Richness$Ar),na.rm=TRUE)
  Richness_mean

  #Fis
  Fis <- t(sapply(seppop(genind_data), function(ls) basic.stats(ls)$perloc$Fis))
  Fis.pop <- rowMeans(as.matrix(Fis),na.rm=TRUE)

  #Store in a dataframe
  Div.stats.1dataset <- data.frame(Pop = names(Hobs.pop),
                        n = Sum$n.by.pop,
                        Hobs = Hobs.pop,
                        Hexp = Hexp.pop,
                        Ar = Richness_mean,
                        Fis = Fis.pop)
  Div.stats.1dataset$Dataset <- genind_data_name
  Div.stats.1dataset$Marker_type <- marker_type

  #Keep all in a common dataframe
  Summary_div_stats <- rbind(Summary_div_stats,
                           Div.stats.1dataset)
  
  #cleaning
  rm(Div.stats.1dataset,Fis,Fis.pop,genind_data,genind_data_name,Hexp,Hexp.pop,Hobs,Hobs.pop,Richness,Richness_mean,Sum)
}
```

Add by hand Antirrhinum charidemi dataset (I do it separately because there is 1 pop that has one locus less than the 2 others)

```{r}
i="FINAL_datasets/microsat/genind_Antirrhinum_charidemi_HeuertzForrest_2024.Rdata"

load(i)
  genind_data_name <- sub("\\.Rdata$", "", basename(i))
  genind_data <- get(genind_data_name)
  
  Sum <- summary(genind_data)

  #Heterozygosity
  Hobs <- sapply(seppop(genind_data,drop=TRUE), function(ls) summary(ls)$Hobs)
  Hexp <- sapply(seppop(genind_data,drop=TRUE), function(ls) summary(ls)$Hexp)

  Hobs.pop <- lapply(Hobs, mean, na.rm=TRUE)
   Hexp.pop <- lapply(Hexp, mean, na.rm=TRUE)
   
  #Allelic Richness
  Richness <- hierfstat::allelic.richness(genind_data, diploid=TRUE)

  Richness_mean <- colMeans(as.matrix(Richness$Ar),na.rm=TRUE)
  Richness_mean

  #Fis
  Fis <- t(sapply(seppop(genind_data), function(ls) basic.stats(ls)$perloc$Fis))
  Fis.pop <- rowMeans(as.matrix(Fis),na.rm=TRUE)

  #Store in a dataframe
  Div.stats.1dataset <- data.frame(Pop = names(Hobs.pop),
                        n = Sum$n.by.pop,
                        Hobs = unlist(Hobs.pop),
                        Hexp = unlist(Hexp.pop),
                        Ar = Richness_mean,
                        Fis = Fis.pop)
  Div.stats.1dataset$Dataset <- genind_data_name
  Div.stats.1dataset$Marker_type <- marker_type
  
    #Keep all in a common dataframe
  Summary_div_stats <- rbind(Summary_div_stats,
                           Div.stats.1dataset)
  
  #cleaning
  rm(Div.stats.1dataset,Fis,Fis.pop,genind_data,genind_data_name,Hexp,Hexp.pop,Hobs,Hobs.pop,Richness,Richness_mean,Sum)
```


Make sure to remove the remaining pop with less than 20 ind. 

```{r}
Summary_div_stats <- Summary_div_stats %>%
                        filter(n >= 20)
#or use a if in the loop earlier
```

Save the file

```{r}
write.csv(Summary_div_stats, file="FINAL_datasets/microsat_summary_div_stats.csv")
```


# Pairwise values for differentiation stats

## Get the pairwise values

Let's save everything in a list ()
```{r}
microsat_files_list <- list.files("FINAL_datasets/microsat/", pattern = "\\.Rdata$", full.names = TRUE)
Pairwise_values_by_dataset <- list()

for (i in microsat_files_list){
  genind_data_name <- sub("\\.Rdata$", "", basename(i)) #extract the name of the dataset
  genind_data <- get(genind_data_name) #assign the studied dataset to a single name

  # Pairwise values
  fst <- genet.dist(genind_data, diploid=TRUE, method="WC84")
  gst_pr_nei <- pairwise_Gst_Nei(genind_data)
  jost_D <- pairwise_D(genind_data)

  # Keep the pairwise matrices in a list
  Pairwise_matrices_stats_diff <- list("fst" = fst,
                                     "gst_pr_nei" = gst_pr_nei,
                                     "jost_D" = jost_D)
  # Save that list in a list to keep data for all datasets
  Pairwise_values_by_dataset[[genind_data_name]] <- Pairwise_matrices_stats_diff
  
  # Cleaning
  rm(genind_data, genind_data_name, fst, gst_pr_nei, jost_D, Pairwise_matrices_stats_diff)
}
```

## TO ADD OR MODIFY A dataset
```{r}
load("FINAL_datasets/microsat_Pairwise_values_by_dataset.Rdata")
```


```{r}
#remaining_ssrs_files <- c("FINAL_datasets/microsat/genind_Ursus_arctos_Kopatz_2021.Rdata")
remaining_ssrs_files <- c("FINAL_datasets/microsat/genind_Coenagrion_mercuriale_Leveque_2024.Rdata")

for (i in remaining_ssrs_files){
  load(i)#load dataset
  print(i)
  genind_data_name <- sub("\\.Rdata$", "", basename(i)) #extract the name of the dataset
  dataset_all_pop <- get(genind_data_name) #assign the studied dataset to a single name
  
  # Pairwise values
  fst <- genet.dist(dataset_all_pop, diploid=TRUE, method="WC84")
  
  #from mmod package
  gst_pr_nei <- pairwise_Gst_Nei(dataset_all_pop)
  jost_D <- pairwise_D(dataset_all_pop)

  # Keep the pairwise matrices in a list
  Pairwise_matrices_stats_diff <- list(
                                      "fst" = fst,#,
                                     "gst_pr_nei" = gst_pr_nei,#,
                                     "jost_D" = jost_D)
  
  # Save that list in a list to keep data for all datasets
  Pairwise_values_by_dataset_microsat[[genind_data_name]] <- Pairwise_matrices_stats_diff
  
  # Cleaning
  rm(dataset_all_pop, genind_data_name, jost_D, gst_pr_nei, fst, Pairwise_matrices_stats_diff)
}
```
Save all
```{r}
save(Pairwise_values_by_dataset_microsat,file="FINAL_datasets/microsat_Pairwise_values_by_dataset.Rdata")
```

## Average pairwise values for a given pop

```{r}
load("FINAL_datasets/microsat_Pairwise_values_by_dataset.Rdata")
mean_Pairwise_values_by_pop_microsat <- matrix(ncol=4, nrow=0)

for (i in names(Pairwise_values_by_dataset_microsat)){
  stats_diff <- c("fst","gst_pr_nei","jost_D")
  for (measure in stats_diff){
    
    labels <- attr(Pairwise_values_by_dataset_microsat[[i]][[measure]], "Labels")
    values <- as.matrix(Pairwise_values_by_dataset_microsat[[i]][[measure]])
    
    for (pop in labels){
      all_pairwise_1pop <- values[pop, ]
      all_pairwise_1pop <- all_pairwise_1pop[names(all_pairwise_1pop) != pop] #to remove pop itself
      
      mean_Pairwise_values_by_pop_microsat <- rbind(mean_Pairwise_values_by_pop_microsat,
                                           c(Dataset = i,
                                              Diff_stat = measure, 
                                              Pop = pop,
                                              mean_pairwise = mean(all_pairwise_1pop)))
    }
    
  }
}

mean_Pairwise_values_by_pop_microsat <- as.data.frame(mean_Pairwise_values_by_pop_microsat)
mean_Pairwise_values_by_pop_microsat$mean_pairwise <- as.numeric(mean_Pairwise_values_by_pop_microsat$mean_pairwise)

mean_Pairwise_values_by_pop_microsat <- mean_Pairwise_values_by_pop_microsat %>%
                                                pivot_wider(names_from = Diff_stat, values_from = mean_pairwise)
```

Add this to the summary diversity statistics 

```{r}
Summary_div_diff_stats_microsat <- Summary_div_stats %>% left_join(mean_Pairwise_values_by_pop_microsat %>% select(-Dataset), by="Pop")
```


# Save the data

```{r}
#cleaning
rm(list=ls()[startsWith(ls(), "genind_")])

#renaming
Pairwise_values_by_dataset_microsat <- Pairwise_values_by_dataset

#saving
save(Pairwise_values_by_dataset_microsat, file="FINAL_datasets/microsat_Pairwise_values_by_dataset.Rdata")
write.table(Summary_div_diff_stats_microsat, file="FINAL_datasets/microsat_summary_div_diff_stats_by_pop.csv")

```



