---
title: "Calculate Bayescan Fst"
author: "MGH"
date: "2025-03-13"
output: html_document
---

Here the point of this code is to prepare the files to run Bayescan (that will be done on the cluster though!)

So my idea is to get genepop files from my genind objects containing full information of the different datasets and then use a .sh code with the PGDSpider pilot to transform these Genepop files into Bayescan format files. 
Here, we need one Genepop file with all the studied population of a given dataset. 

Then I'll put everything on the cluster and run the code. 

Finally, after downloading the result files on the cluster, I need a code to gather everything automatically in a single table of result. 

# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  eval=FALSE) 
```

```{r}
rm(list=ls(all=TRUE))
#!!!!! Change directory according your own directory !!!!!
#setwd()
```

```{r}
library(graph4lg)
library(adegenet)
library(dartR)
library(RLDNe)

library(dplyr)
library(ggplot2)
library(tidyverse)
```

# 1. Get a genepop file from an R genind object

## SNPs

This has already beeen done, check in the file 3.Filt_pop_assignation_transfo_genepop.Rmd

## SSR

Select the files to convert into genepop
```{r}
full_microsat_files <- list.files("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v2/FINAL_datasets/microsat", full.names = TRUE)

#if has already been done for some datasets
#full_microsat_files <- full_microsat_files[-c(1:6)]
#full_microsat_files
```

Convert from genind object in R to Genepop files (outside of R)
```{r}
for (i in full_microsat_files){
  load(i)
  
  genind_data_name <- sub("\\.Rdata$", "", basename(i))
  genind_data <- get(genind_data_name)
  
  genind_data_name2 <- sub("^genind_","",genind_data_name)
  print(genind_data_name2)
  
  genind_to_genepop(genind_data,output=paste0("genepop_files/microsat/",genind_data_name2,".txt"))
  
}
```

For only one dataset for example
```{r}
i="C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Genetic_data_v2/FINAL_datasets/microsat/genind_Cypripedium_calceolus_Gargiulo_2023.Rdata"
  load(i)
  
  genind_data_name <- sub("\\.Rdata$", "", basename(i))
  genind_data <- get(genind_data_name)
  
  genind_data_name2 <- sub("^genind_","",genind_data_name)
  print(genind_data_name2)
 
#does not work 
genind_to_genepop_2pop(genind_data,output=paste0("genepop_files/microsat/",genind_data_name2,".txt"))
```

Okay so the code does not work here because the dataset has only 2 populations and there is a problem with that in the function !! So I slightly modify it for it to work!

```{r}
genind_to_genepop_2pop <- function(x, output = "data.frame"){

  # Check whether 'x' is a genind object
  if(!inherits(x, "genind")){
    stop("Input 'x' must be an object of class 'genind'.")
  }

  # Get genetic data from x@tab
  data <- x@tab
  # Get pop_names
  pop_names <- x@pop

  ### Return a message if there is only one population
  if(length(unique(pop_names)) == 1){
    message("There is only one population in your dataset")
    unique_pop <- TRUE
  } else {
    unique_pop <- FALSE
  }

  # If the individuals are not ordered by populations, they are reordered
  # with their populations in alphabetic order.
  if(!all(pop_names == rep(pop_names[-which(duplicated(pop_names))],
                           times = table(pop_names)[unique(pop_names)]))){
    message("Individuals in the input data were not ordered, they have
          been ordered by populations and populations in alphabetic order
          for the conversion.")
    pop_names <- as.character(x@pop)[order(as.character(x@pop))]
    data <- x@tab[order(as.character(x@pop)),]
  } else {

    pop_names <- as.character(x@pop)[order(as.character(x@pop))]
    data <- x@tab[order(as.character(x@pop)),]

  }

  ### Identify the type of markers and modify SNPs data for GENEPOP usage
  if(all(unlist(unique(x@all.names)) %in% c("A", "T", "C", "G"))){
    m_type <- "snp"
    message("Your dataset is treated as a SNP dataset.
            Alleles initially coded A, T, C, G were respectively coded
            01, 02, 03 and 04")

    colnames(data) <- gsub(colnames(data),pattern=".A",replacement=".01")
    colnames(data) <- gsub(colnames(data),pattern=".T",replacement=".02")
    colnames(data) <- gsub(colnames(data),pattern=".C",replacement=".03")
    colnames(data) <- gsub(colnames(data),pattern=".G",replacement=".04")
    #####
  } else {
    m_type <- "msat"
  }

  # Get loci names
  loci_names_l <- x@loc.fac
  # Create an empty data.frame
  loc_all <- data.frame(col = colnames(data))
  # Check whether the names of the locus.allele combination are well formatted
  # with a '.' between the locus and the allele names.
  if(all(stringr::str_count(colnames(data), "\\.") == 1) != TRUE){
    stop("The columns' names of x@tab must be of form 'locus.allele' with only 1
         '.' between locus and allele")
  }
  # Separate the locus and the allele
  loc_all <- tidyr::separate(loc_all, col = 1, sep = "\\.",
                             into = c("locus", "allele"))

  # Avoid problems when allele names have not the same number of characters
  max_all_chr <- max(nchar(loc_all$allele))
  for(i in 1:nrow(loc_all)){
    name_all_chr <- nchar(loc_all[i, "allele"])
    loc_all[i, "allele"] <- ifelse(name_all_chr < max_all_chr,
                                   paste0(rep("0", times = max_all_chr - name_all_chr),
                                          loc_all[i, "allele"]),
                                   loc_all[i, "allele"])
  }


  loci_names <- as.character(loci_names_l[-which(duplicated(loci_names_l))])
  n.loci <- length(loci_names_l[-which(duplicated(loci_names_l))]  )

  # Create the data.frame data_gpop which will contain the genetic data
  # The first column contains the ID of each individual
  data_gpop <- data.frame(id = paste(pop_names, "_", row.names(data), ",",
                                     sep = ""))

  data_gpop[] <- lapply(data_gpop, gsub, pattern = " ", replacement = "")

  # The loop will run over all the loci
  for (i in 1:n.loci){
    # loc is the locus considered
    loc <- loci_names[i]
    # a is an empty vector which will contain the two alleles coding of all
    # individuals over all the loci
    a <- c()
    # The loop will run over all the individuals
    for (j in 1:nrow(data)){
      # Get the column corresponding to the locus
      col_loc <- which(loc_all[, 'locus'] == loc)
      # Get the locus.allele at which individual j is homozygote or heterozygote
      hom <- which(data[j,col_loc] == 2)
      het <- which(data[j,col_loc] == 1)

      # If homozygote, then copy the code of the allele twice
      if(length(hom) != 0){
        a[j] <- paste(loc_all[col_loc[hom], 'allele'],
                      loc_all[col_loc[hom], 'allele'],
                      sep = "")
        # If heterozygote, then copy the code of the two alleles of ind j
        # the lower code number is the first
      } else if (length(het) != 0){

        if (as.character(loc_all[col_loc[het[1]], 'allele']) <
            as.character(loc_all[col_loc[het[2]], 'allele'])){

          a[j] <- paste(loc_all[col_loc[het[1]], 'allele'],
                        loc_all[col_loc[het[2]], 'allele'],
                        sep = "")
        } else {
          a[j] <- paste(loc_all[col_loc[het[2]], 'allele'],
                        loc_all[col_loc[het[1]], 'allele'],
                        sep = "")
        }

      } else {
        # If missing data, set 000000 or 0000
        if(nchar(loc_all[1, 'allele'] == 6)){
          a[j] <- "000000"
        } else {
          a[j] <- "0000"
        }
        ###
      }

    }

    # For each locus, add the next column to data_gpop
    data_gpop <- cbind(data_gpop, a)
  }

  # Then, set the colnames
  colnames(data_gpop) <- c("ID", as.character(loci_names))

  # Convert all the columns as characters
  data_gpop[ , ] <- lapply(data_gpop[ , ], as.character)

  # Add the pop_names
  # Distinguish the case where there is only one pop
  if(unique_pop){

    # Create data_gpop2 whose first line only is "POP"
    data_gpop2 <- rbind(c("Pop",rep("", n.loci)), data_gpop)


  } else {


    # Get the rows' numbers of the end of the populations
    num_end_pop <- c()
    for  (i in (1:(length(pop_names) - 1))){
      if (pop_names[i] != pop_names[i + 1]){
        num_end_pop[i] <- i
      } else {
        num_end_pop[i] <- 0
      }
    }
    num_end_pop <- which(num_end_pop != 0)


    # Create data_gpop2 whose first line only is "POP"
    data_gpop2 <- rbind(c("Pop",rep("", n.loci)), data_gpop[1:num_end_pop[1], ])
    
    ############################################################################
    #My personal modif!! 
    # Then, add the rows of each population, one line with "POP", etc..
    #for (i in 1:(length(num_end_pop)-1)){
    #  data_gpop2 <- rbind(data_gpop2, c("Pop", rep("", n.loci)),
    #                      data_gpop[(num_end_pop[i]+1):num_end_pop[i+1], ])
    #}
    ############################################################################
    #So I remove that part because it creates when you have 2 populations 
    # Indeed num_end_pop[i+1] --> NA in that case 
    
    # Add the last population
    data_gpop2 <- rbind(data_gpop2, c("Pop", rep("", n.loci)),
                        data_gpop[(num_end_pop[length(num_end_pop)]+1):nrow(data_gpop), ])

  }

  # The two first lines are the heading and the names of the loci
  if(n.loci > 1){
    data_gpop2 <- rbind(c("Conversion of an object of class 'genind' into a GENEPOP file with the package 'graph4lg'",
                          rep("", n.loci)),
                        c(paste(loci_names[1:(n.loci-1)],",", sep = ""),
                          loci_names[n.loci], ""),
                        data_gpop2)
    colnames(data_gpop2)[2:(length(colnames(data_gpop2))-1)] <-
      paste(colnames(data_gpop2)[2:(length(colnames(data_gpop2))-1)], ",", sep = "")

  } else {
    data_gpop2 <- rbind(c("Conversion of an object of class 'genind' into a GENEPOP file with the package 'graph4lg'",
                          rep("", n.loci)),
                        c(loci_names[n.loci], ""),
                        data_gpop2)
  }

  # Output
  if(output == "data.frame"){
    return(data_gpop2)
  } else if (stringr::str_sub(output,-4,-1) == ".txt") {
    utils::write.table(data_gpop2,file = output, quote=FALSE,
                       row.names=FALSE, col.names=FALSE)
  } else {
    stop("'output' parameter must be 'data.frame' (then the function returns an
object of class data.frame) or a path and file name ending in '.txt'. In the
          latter case, the functions creates a text file in the directory
         specified by the path or in the current working directory when a
         path is not specified.")
  }

}
```



# 2. Get files formated for Bayescan 

Look at code_genepop2_bayescan.sh which uses an executive of PGDSpider to do so. 

# 3. Run Bayescan

Look at Bayescan_for_SSR.sh & Bayescan_for_SNP.sh
Done on the cluster


# 4. Check the results

## SSR

```{r}
microsat_bayescan_list <- list.files("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Fst_bayescan/bayescan_files/microsat/", full.names = TRUE)
microsat_bayescan_list <- sub("*.txt","",basename(microsat_bayescan_list))
microsat_bayescan_list <- sub("^bayescan_","",microsat_bayescan_list)
microsat_bayescan_list
```


```{r}
fst_bayescan_by_pop <- matrix(ncol=6,nrow=0)
colnames(fst_bayescan_by_pop) <- c("Dataset","Marker_type","Pop_index","Ind_from_pop","mean_Fst","median_Fst")
marker_type="microsat"

bayescan_check_algo <- matrix(ncol=6,nrow=0)
colnames(bayescan_check_algo) <- c("Dataset","Marker_type","Nb_iterations_for_fst","Nb_pop","Accept_rate_beta","Accept_rate_ances")

for (i in microsat_bayescan_list){
  
  # Step 1: get the population names back
  file_path <- paste0("genepop_files/microsat/",i,".txt")
  lines <- readLines(file_path)
  pop_indices <- grep("^POP", lines, ignore.case = TRUE)
  pop_names <- lines[pop_indices + 1]
  pop_names <- sub(",.*", "", pop_names)

  # Step 2: get the Fst by pop from Bayescan
  results <- read.csv(paste0("results/microsat/bayescan_",i,".sel"), sep="")
  
  pdf(file = paste0("results/graphics_microsat/Post_distrib_",i,".pdf"))
  
  for (j in 1:(ncol(results)-1)){
    #plot the posterior distribution
    plot(density(results[,j+1]), xlab=pop_names[j],main=paste("Fst",pop_names[j],"posterior distribution"))
    par(new = TRUE)  #to add the hist on the previous graph
    hist(results[, j+1], col = rgb(1, 0, 0, 0.5), border = "white", 
          xlab = "", ylab = "", main = "", axes = FALSE)

    #check convergence was okay
    plot(results[,j+1], xlab=pop_names[j],main=paste("MCM Chain",pop_names[j]))
    
    #main info in a tab
    fst_bayescan_by_pop <- rbind(fst_bayescan_by_pop,
                             c(i,marker_type,j,pop_names[j],
                               mean(results[,j+1]),
                               median(results[,j+1])))
  }
  dev.off()
  
  
  # Check the algorithm went well
  acrte <- read.csv(paste0("results/microsat/bayescan_",i,"_AccRte.txt"), sep="")
  
  beta_acrte <- mean(acrte$beta)
  ances_acrte <- mean(acrte$ances)
  nb_iter <- dim(results)[1]
  nb_pop <- dim(results)[2]-1
  bayescan_check_algo <- rbind(bayescan_check_algo,
                               c(i,marker_type,nb_iter,nb_pop,beta_acrte,ances_acrte))
  
  
  #cleaning
  rm(nb_iter,nb_pop,beta_acrte,ances_acrte)
}
```

Okay so I checked the distribution plots --> it all seems beautiful! So I will use the mean Fst for all the populations

With the acceptance rate, it seems okay, some are below the 25%. 

For the number of iterations, here we have one of 10 iterations (it's from the settings) --> so with 5000 here, it means that we have 50000 in total

To add the Fst to my analyses, I will have to take back the population name from the individual name I got. I will do so by hand to save time, using the column with the name of first individual of the corresponding population.

```{r}
write.csv(bayescan_check_algo, file="results/microsat_check_Bayescan_algo.csv")
write.csv(fst_bayescan_by_pop, file="results/microsat_results_Fst_Bayescan.csv")
```


## SNPs

```{r}
snps_bayescan_list <- list.files("C:/Users/mgharribey/Documents/These_BIOGECO/Article1/analyses_article1/Summary_stats_genetic/Fst_bayescan/bayescan_files/1000snps_corrected/", full.names = TRUE)
snps_bayescan_list <- sub("*.txt","",basename(snps_bayescan_list))
snps_bayescan_list <- sub("^bayescan_","",snps_bayescan_list)
snps_bayescan_list
```


Get back the name of populations (assume same order as in the genepop files)
```{r}
fst_bayescan_by_pop <- matrix(ncol=6,nrow=0)
colnames(fst_bayescan_by_pop) <- c("Dataset","Marker_type","Pop_index","Ind_from_pop","mean_Fst","median_Fst")
marker_type="snps"

bayescan_check_algo <- matrix(ncol=6,nrow=0)
colnames(bayescan_check_algo) <- c("Dataset","Marker_type","Nb_iterations_for_fst","Nb_pop","Accept_rate_beta","Accept_rate_ances")

for (i in snps_bayescan_list){
  
  # Step 1: get the population names back
  file_path <- paste0("genepop_files/1000snps_corrected/",i,".txt")
  lines <- readLines(file_path)
  pop_indices <- grep("^POP", lines, ignore.case = TRUE)
  pop_names <- lines[pop_indices + 1]
  pop_names <- sub(",.*", "", pop_names)

  # Step 2: get the Fst by pop from Bayescan
  results <- read.csv(paste0("results/snps/bayescan_",i,".sel"), sep="")
  
  pdf(file = paste0("results/graphics_snps/Post_distrib_",i,".pdf"))
  
  for (j in 1:(ncol(results)-1)){
    #plot the posterior distribution
    plot(density(results[,j+1]), xlab=pop_names[j],main=paste("Fst",pop_names[j],"posterior distribution"))
    par(new = TRUE)  #to add the hist on the previous graph
    hist(results[, j+1], col = rgb(1, 0, 0, 0.5), border = "white", 
          xlab = "", ylab = "", main = "", axes = FALSE)

    #check convergence was okay
    plot(results[,j+1], xlab=pop_names[j],main=paste("MCM Chain",pop_names[j]))
    
    #main info in a tab
    fst_bayescan_by_pop <- rbind(fst_bayescan_by_pop,
                             c(i,marker_type,j,pop_names[j],
                               mean(results[,j+1]),
                               median(results[,j+1])))
  }
  dev.off()
  
  
  # Check the algorithm went well
  acrte <- read.csv(paste0("results/snps/bayescan_",i,"_AccRte.txt"), sep="")
  
  beta_acrte <- mean(acrte$beta)
  ances_acrte <- mean(acrte$ances)
  nb_iter <- dim(results)[1]
  nb_pop <- dim(results)[2]-1
  bayescan_check_algo <- rbind(bayescan_check_algo,
                               c(i,marker_type,nb_iter,nb_pop,beta_acrte,ances_acrte))
  
  
  #cleaning
  rm(nb_iter,nb_pop,beta_acrte,ances_acrte)
}
```

Checked the distribution plots --> in the vast majority of populations, distribution of values seems normal. So I can use the mean Fst for all the populations. 

For the Markov Chains, seem good too, with a good repartition of points, mean the chains converge well

The acceptance rates are between 0.2 and 0.25 in a lot cases (both for beta and alpha)... Foll & Gaggiotti suggest to have between 0.25 and 0.45 acceptance rates, but not a lot of details in the bibliography. In the case of low acceptance rates, it would mean that most moves are rejected because they correspond to areas of low posterior probability. 

Don't forget to add by hand the name of populations, using the first individual name of each population. 
